\section{Motivation and Research Questions}
\label{sec:llm-motivation}

Chapter~\ref{sec:continued-pretraining} established that a specialized RoBERTa encoder, augmented with metalinguistic auxiliary objectives, reaches 62.19 F1-Macro on AllSidesV2, a small but real improvement over the POLITICS contrastive baseline. That result raises an immediate practical question. LLM APIs are increasingly the default infrastructure for downstream NLP work: they offer zero-shot capability out of the box, low engineering overhead, and broad accessibility for practitioners who do not have the resources to run continued pre-training. If a general-purpose LLM, fine-tuned on the same task data, can match or exceed our specialized continued-pretraining pipeline at meaningfully lower engineering cost, then the right recommendation for downstream users of bias detection changes substantially. Conversely, if it cannot, that itself is an important negative result about the limits of fine-tuning general-purpose LLMs on nuanced directional tasks. Either answer matters for the practical viability of LLM-based content moderation and bias-detection deployments, where data efficiency, input length sensitivity, and scaling behavior directly determine whether a system can be operated at all.

This is the pivot that motivates the current chapter: having shown what specialized continued pre-training can do, we now ask what general-purpose LLM fine-tuning can do on the same task and the same data. While our baseline experiments (Chapter~\ref{sec:baselines}) demonstrated that zero-shot large language models achieve competitive but suboptimal performance compared to domain-adapted transformers, the potential for task-specific fine-tuning to close this performance gap remains underexplored. The superior performance of the specialized POLITICS model over general-purpose LLMs raises several important questions about data efficiency (how much task-specific training data is required for LLMs to achieve competitive performance with domain-adapted transformers), input length sensitivity (whether LLMs exhibit different performance characteristics when fine-tuned on truncated versus full-length news articles), and scalability properties (the relationship between training data volume and classification performance for political bias detection). These questions are particularly relevant given the substantial computational and economic costs associated with LLM fine-tuning, necessitating careful analysis of cost-benefit trade-offs in practical deployment scenarios.

