\section{Experimental Results and Analysis}
\label{sec:llm-results}

Table~\ref{tab:llm_detailed_results} presents comprehensive results from our fine-tuning experiments across all evaluated configurations. The results reveal several important patterns regarding LLM behavior in political bias classification tasks.

\subsection{Training Data Scaling Effects}

The relationship between training set size and model performance exhibits non-monotonic behavior that varies significantly between full-text and truncated input conditions. For full-text experiments, we observe high performance variability with small training sets: F1-Macro scores range from 25.04 to 33.37 with 150--300 samples, indicating insufficient supervision for stable learning. Performance stabilizes substantially with 1,000 training samples (F1-Macro = 69.23), achieving near-optimal results that improve only marginally with additional data (72.48 with 2,000 samples).

This pattern suggests that GPT-4o-mini requires a minimum threshold of approximately 1,000 examples to achieve reliable fine-tuning for political bias classification when processing full-length articles. The minimal improvement between 1,000 and 2,000 samples (0.14 F1 points) indicates diminishing returns, suggesting that the model approaches its performance ceiling for this task configuration.

The 300-sample full-text result (F1 = 25.04) is substantially worse than the equivalent zero-shot baseline (51.86) and is too large a gap to attribute to ordinary scaling noise. The most plausible mechanism is representation overwriting: with very few examples and very long inputs, each gradient update touches a small region of the input space but applies a large effective parameter change, and the optimizer pulls the model away from its general-purpose competence on long political prose without providing enough supervised signal to install a coherent task-specific behavior in its place. The model ends up worse than its untouched zero-shot self because fine-tuning has degraded what it already knew without giving it anything stable to replace it with. The comparatively well-behaved truncated-input results in the same low-data regime support this reading: short inputs reduce the per-example footprint of each update and make small-data fine-tuning safer.

\subsection{Input Length Sensitivity Analysis}

The comparison between full-text and 400-word truncation reveals markedly different performance characteristics. Truncated inputs demonstrate more consistent fine-tuning improvements across all training set sizes, with every configuration outperforming the corresponding zero-shot baseline. Most notably, truncated inputs achieve ``competitive'' performance even with limited supervision: the 150-sample configuration yields F1 = 62.88, which beats the equivalent 150-sample full-text result (33.37) and the zero-shot baseline (51.86), and exceeds the POLITICS contrastive baseline (57.66). It is, however, well below what GPT-4o-mini reaches on full text given enough data (69.23 at 1,000 samples), so ``competitive'' here means ``strong relative to short-input baselines'' rather than ``equivalent to a well-fitted full-text model.'' Full-text fine-tuning appears to require substantially more examples before its scaling kicks in.

The superior performance on shorter texts suggests that LLMs may be optimized for processing concise, focused content rather than long-form articles with potential noise and irrelevant information.

\subsection{Performance Comparison with Specialized Models}

The best fine-tuned GPT-4o-mini configuration (72.48 F1-Macro with 2,000 full-text samples) substantially exceeds our human baseline (41.94) and exceeds the performance of the specialized POLITICS model (57.66 on the same evaluation set). This comparison must be interpreted carefully on several grounds. The fine-tuned LLM requires explicit task-specific supervision while the POLITICS model leverages contrastive pre-training without direct ideology labels, which is a real cost-of-training asymmetry in POLITICS's favor. More importantly, GPT-4o-mini was pre-trained on web-scale corpora that very plausibly include AllSides articles, AllSides-derived datasets, and adjacent political-bias commentary, so part of the headline F1 may reflect prior exposure to the labels rather than learned content reasoning, a contamination concern that does not apply to POLITICS in any comparable form. The fine-tuned-LLM-vs-POLITICS comparison is therefore not apples-to-apples, and the causal-probing analysis in Chapter~\ref{sec:causal-inference} is the diagnostic that lets us interpret whether the F1 gap reflects genuine ideological reasoning or shortcut learning.

We fine-tuned only GPT-4o-mini, not o3-mini, GPT-5-nano, or any LLaMA variant, in this chapter. The reason is straightforward and worth stating explicitly: the cost of running a controlled fine-tuning sweep across multiple training set sizes and input lengths on a frontier-tier API model was the binding constraint on what we could afford, and GPT-4o-mini was the most competitive model we could fine-tune within the available budget. A cross-model fine-tuning ablation would meaningfully strengthen the contribution and is flagged as future work.

The competitive performance of fine-tuned LLMs validates their potential for political bias classification when sufficient training data is available. However, the requirement for 1,000+ examples and the associated computational costs may limit practical applicability compared to specialized architectures that achieve strong performance through unsupervised or self-supervised learning approaches.
