\section{Future Research Directions}
\label{sec:future-directions}

Future research should investigate additional metalinguistic dimensions that may provide complementary supervision signals for political bias detection. Syntactic complexity measures are one promising candidate, as political writing styles may exhibit systematic differences in sentence length distribution and grammatical structure patterns that correlate with ideological orientation. Discourse markers and argumentation patterns---the use of specific rhetorical devices, evidence types, and argumentation strategies---may also provide discriminative signals for ideological classification. Named entity coverage analysis, examining systematic differences in which political figures, organizations, or geographic locations receive coverage, could capture ideological source preferences that complement our thematic and tonal features.

Advanced multi-task learning architectures could address the limitations of our current static approach. As a concrete mechanism: a cross-task attention layer between the theme head and the tone head would let the model condition tonal interpretation on detected topical structure, so that if the theme head identifies that an article is about immigration, attention can up-weight sentiment signals on immigration-related entities relative to unrelated sentiment dimensions, instead of treating all sentiment dimensions independently as the current architecture does. The current additive multi-task setup forces auxiliary signals to interact only through the shared encoder, whereas an explicit cross-task attention layer would let them interact at the head level. Hierarchical task organization, structuring auxiliary tasks to reflect theoretical relationships between metalinguistic dimensions, could similarly improve learning efficiency and interpretability.

The causal analysis framework presented in Chapter~\ref{sec:causal-inference} opens several additional research directions. Our current framework treats each topic sentiment as a separate treatment or mediator; an alternative co-treatment formulation would model multiple topic sentiments jointly, estimating their combined causal effect on ideology classification and capturing interaction effects that the current single-treatment design may miss. The mediation-as-diagnostic framework could be applied beyond political ideology to other annotation tasks where LLM labels are used as silver-standard training data, such as sentiment analysis, stance detection, or toxicity classification. Comparing mediation pathways across human and LLM annotators in these domains would determine whether the sensitivity patterns we find are specific to ideology or reflect a more general tendency of LLMs.

The directions above span very different scopes and effort levels, and it is more useful to triage them than to present them as a flat list. We group them below in order of increasing scope, pairing each future-work item with the limitation it most directly addresses.

The most immediate follow-up is broader model coverage in the causal analysis: extending the framework of Chapter~\ref{sec:causal-inference} to additional LLM families and to models with reported ideological tendencies such as Grok \citep{xai2024grok} would address the LLM-family-coverage gap flagged in the methodological limitations and could be completed without substantial methodological development.

A second immediate extension is evaluating chain-of-thought (COT) prompting on causal metrics rather than classification accuracy alone. The OPRO analysis in Appendix~\ref{appendix:opro} found that COT prompts produced slightly lower F1 than the baseline prompt (51.47 vs.\ 51.86), but F1 measures classification accuracy, not the causal fidelity of the resulting labels. The shortcut learning evidence from Community~1 (Section~\ref{sec:shortcut-learning}) suggests that models with comparable F1 can produce very different causal effect patterns depending on which textual features drive their predictions. \citet{yuan2024llmsovercomeshortcutlearning} and \citet{sun-etal-2024-exploring} identify COT reasoning as particularly effective at mitigating shortcut learning by forcing explicit intermediate reasoning steps. COT-generated labels could therefore produce ATE estimates that are closer to the human baseline if the reasoning process helps the model attend to genuine ideological signals rather than surface correlations. Evaluating COT labels through the causal framework of Chapter~\ref{sec:causal-inference} would determine whether explicit reasoning at inference time serves as an alternative to fine-tuning for aligning causal effect estimates with human judgments.

A second set of extensions would more substantively broaden the contribution of the line of research initiated in this thesis. Co-treatment causal models that estimate the joint effect of multiple topic sentiments simultaneously, along with their interaction effects, would address the single-treatment limitation of the current mediation framework. Controlled sentiment-input experiments that hold the ideology classifier fixed while varying the sentiment extractor, and vice versa, would help disentangle the two candidate mechanisms behind Llama-3.3-70B's attenuation pattern identified in Section~\ref{sec:community-level-results}. Cross-task attention architectures along the lines described above would address the additive-loss limitation of the current multi-task setup.

Finally, the mediation-as-diagnostic framework is the most portable artifact of this thesis, and generalizing it beyond ideology would carry the methodological contribution into other domains. Future work could apply the framework to other annotation tasks where LLM labels are used as silver-standard data, such as sentiment analysis, stance detection, and toxicity classification, in order to determine whether the shortcut-learning patterns we report here are specific to political ideology or reflect a more general property of LLM annotators.

