\section{Methodological Limitations and Constraints}
\label{sec:methodological-limitations}

\subsection{Data Coverage and Metadata Availability}

A fundamental limitation of our metalinguistic feature integration approach stems from incomplete metadata coverage across our training corpus. The effectiveness of our multi-task learning framework was substantially constrained by the availability of GDELT annotations: only 1.6 million articles (35\%) from our 4.5 million article \textsc{BigNewsBLN} corpus contained the requisite \textsc{V2Themes} and \textsc{V2Tone} metadata necessary for auxiliary supervision.

This data sparsity arose primarily from temporal coverage gaps, as many articles in our dataset were published before the GDELT Project began comprehensive annotation collection. The resulting imbalanced supervision likely diminished the potential impact of our multi-task training approach, as auxiliary objectives received training signals for only a subset of examples while the primary triplet-loss objective operated on the complete corpus. The Table~\ref{tab:pretraining-combined} subset-vs-full comparison in Chapter~\ref{sec:continued-pretraining} gives a concrete handle on the magnitude of this loss: on the 1.6M-article subset where every article carries GDELT metadata, the best metalinguistic configurations are roughly competitive with the pure-contrastive baseline, whereas on the full 4.5M-article corpus (35\% GDELT coverage) the same configurations gain several F1 points above the pure-contrastive baseline. As a counterfactual extrapolation (emphatically not a measurement), if the per-example contribution of complete GDELT supervision held linearly across the full corpus, an idealized 100\%-coverage version of our pre-training pipeline could be expected to recover an additional several F1-Macro points beyond the 62.19 we report. We treat this number as a directional estimate of the data-coverage ceiling, not as a prediction.

The temporal bias in metadata availability also introduces potential confounding effects, as articles with GDELT annotations may systematically differ from those without in ways that correlate with ideological expression patterns. More recent articles might reflect different political discourse styles, topic priorities, or source characteristics compared to historical content, potentially biasing our auxiliary task learning toward contemporary linguistic patterns.

\subsection{Evaluation Dataset Constraints}

Our empirical evaluation relies exclusively on the AllSides dataset, which, while providing high-quality article-level annotations, constrains the scope of our findings to a single annotation scheme and political taxonomy. The AllSides left-center-right framework represents one particular operationalization of political ideology that may not capture the full complexity of ideological expression in contemporary media discourse.

Additionally, the AllSides dataset focuses exclusively on mainstream news sources with established reputations, and the resulting scoping bias has two specific implications for how our results should be read. First, mainstream outlets tend to use more formal, standardized editorial language, which makes ideology classification structurally \emph{easier} than it would be on partisan, alternative, or experimental publications: the linguistic features our models exploit (theme, tone, framing) are most legible when they sit on top of a consistent editorial register. Performance numbers in this thesis should therefore be read as upper bounds on what the same approach would achieve in noisier media environments. Second, fringe and non-mainstream media may exhibit ideological dimensions that are orthogonal to the left-center-right axis altogether (libertarian-versus-authoritarian, populist-versus-establishment, anti-system-versus-system), and the AllSides taxonomy by construction collapses these dimensions onto a single horizontal spectrum that this thesis cannot capture. The source selection criteria employed by AllSides therefore introduce systematic biases both toward particular types of content and toward a particular axis of ideological variation.

\subsection{Cultural and Linguistic Scope}

Our investigation is fundamentally limited to English-language content from U.S. news sources, constraining applicability to international political systems and multilingual contexts. Political discourse patterns, media landscape structures, and ideological frameworks vary substantially across national contexts, potentially limiting the transferability of our metalinguistic feature integration approach.

The left-center-right ideological spectrum that underlies our classification framework may inadequately represent political dimensions in multi-party parliamentary systems, contexts with ethnic or religious political divisions, or societies with fundamentally different ideological cleavages. For example, European political systems often feature distinct liberal, social democratic, conservative, and green party traditions that may not map cleanly onto the American left-right continuum.

Cross-linguistic generalization presents additional challenges, but it is worth distinguishing precisely which part of the pipeline the linguistic-scope limitation actually constrains. The constraint applies most directly to metalinguistic feature \emph{extraction}: GDELT's V2Themes and V2Tone lexicons are English-tuned, sentiment and thematic categorization schemes that prove effective for English may require substantial adaptation for morphologically rich languages, tonal languages, or languages with different information-structure conventions, and the auxiliary supervision we rely on in Chapter~\ref{sec:continued-pretraining} is therefore not automatically portable to non-English corpora. By contrast, the causal-analysis framework in Chapter~\ref{sec:causal-inference} is essentially language-agnostic: given any pipeline that produces ideology labels and topic-level sentiment scores (in any language), the DML and mediation machinery can be applied without modification. Cross-linguistic extension is therefore best framed as a feature-extraction problem rather than a framework limitation: the metalinguistic objectives need new lexicons, but the causal probing built on top of them does not.

\subsection{Causal Analysis Limitations}

The causal analysis in Chapter~\ref{sec:causal-inference} introduces several additional constraints. First, our sentiment annotations are generated by an LLM (Llama-3.3-70B), while the study investigates differences between human and LLM ideology annotations. If the sentiment extractor itself has systematic biases, the causal analysis may be partly confounded by shared model biases between the sentiment extraction and LLM ideology annotation stages. We mitigate this risk by using distinct model sessions for sentiment extraction and ideology prediction, but the possibility of correlated systematic biases between the two stages cannot be fully ruled out.

Second, LLM-generated sentiment annotations may not perfectly capture the nuanced sentiment expressed in news articles. Topic-level sentiment aggregation further compresses this signal, potentially obscuring within-topic variation. These measurement limitations could attenuate causal estimates and affect the validity of null findings.

Third, our causal framework controls for topic presence but does not account for outlet identity, author, or publication date. Different outlets may have systematic sentiment patterns that confound the sentiment-ideology relationship. While controlling for outlet identity would address this, it would also substantially reduce the variation available for estimation given the correlation between outlet and ideology.

Fourth, with 11,516 articles distributed across topics and communities, some analyses operate on small effective sample sizes (as few as 103 articles per community after topic filtering). This limits statistical power, particularly for the community-level and mediation analyses. The null findings for human annotations should therefore be interpreted with caution, as they may reflect insufficient power rather than true absence of effects.
