\section{Causal Analysis}
\label{sec:causal-discussion}

The causal analysis in Chapter~\ref{sec:causal-inference} reveals that topic-level sentiment carries a robust causal relationship with ideology classification, but only for inherently polarising topics. Seven of 20 qualifying topics produce significant effects that replicate across all or nearly all annotation paradigms (Table~\ref{tab:multi-continuous}), and all five paradigms agree on a positive Total Effect for the Donald Trump mediation pathway (Table~\ref{tab:te-trump}). This broad cross-paradigm agreement on partisan topics contrasts with substantial annotator disagreement on non-partisan topics (Table~\ref{tab:multi-nonsig}), and with the community-level results, where Community~2 (Public Health) achieves full cross-paradigm replication but the remaining significant communities show more paradigm-specific patterns (Section~\ref{sec:community-level-results}). As only the ideology labels vary across the five datasets, these patterns reflect systematic differences in how each labelling paradigm assigns ideological stance to article text.

\subsection{Polarisation and Causal Signal}

The most striking finding is that effect significance does not track article count but instead tracks the polarising quality of the topic. The three largest topics by article count (Elections, Politics, Presidential Elections) produce no significant effects under any paradigm, while smaller but explicitly partisan topics (Democratic Party, Hillary Clinton, Republican Party) are significant across all five. The directional structure of the significant effects is consistent with partisan expectations: topics associated with Democratic figures and positions (Hillary Clinton, Democratic Party, Barack Obama, Joe Biden, Healthcare) carry uniformly negative ATEs, meaning that more positive sentiment toward these topics shifts ideology predictions leftward. Topics associated with Republican figures (Donald Trump, Republican Party) carry uniformly positive ATEs. This alignment holds across all five annotation paradigms, suggesting that the causal signal reflects genuine partisan structure in the corpus rather than annotation artifacts.

\subsubsection{Russia as an Outlier}

Russia is a notable exception to the pattern of partisan alignment. Despite covering a geopolitically salient topic ($N = 270$), Russia produces the most divergent directional profile of any topic in Table~\ref{tab:multi-nonsig}: Fine-tuned RoBERTa ($+0.0044$), Fine-tuned GPT-4o-mini ($+0.0036$), and Llama-3.3-70B ($+0.0144$) estimate positive effects, while Human ($-0.0008$) and GPT-4o-mini baseline ($-0.0044$) estimate negative effects. All effect sizes are close to zero and non-significant, but the directional split is informative.

\citet{smirnova2017ideology} document a shift toward more negative sentiment in New York Times coverage of Russia following the onset of the Russia-Ukraine conflict, noting that this sentiment shift was concentrated in a left-leaning publication. Our corpus, which spans a similar period and draws from outlets across the ideological spectrum, may reflect a version of this dynamic: Russia-related sentiment does not map cleanly onto the left-right axis because the topic cuts across traditional partisan lines. The directional split between annotation paradigms is itself instructive. GPT-4o-mini baseline and Human annotations both estimate a (non-significant) negative relationship between Russia sentiment and ideology, while fine-tuned GPT-4o-mini estimates a positive one. This divergence between GPT-4o-mini baseline and its fine-tuned counterpart, which share the same architecture but differ only in task-specific training, suggests that fine-tuning may introduce or amplify a directional association that is absent in the zero-shot model. Whether this reflects a genuine pattern in the fine-tuning data or a spurious association is an open question and a candidate for future investigation with targeted ablation studies.

\subsection{Effect Size Comparison Across Paradigms}

Section~\ref{sec:effect-magnitudes} established that ATEs are uniformly small across both analysis levels, positioning topic-specific sentiment as a supplementary rather than dominant signal in ideology classification. Within this regime of small effects, a consistent magnitude ordering nonetheless emerges across the seven topics with cross-paradigm significance in Table~\ref{tab:multi-continuous}: Llama-3.3-70B and GPT-4o-mini (baseline) produce the largest ATEs, while Human annotations produce the smallest. Fine-tuned RoBERTa and Fine-tuned GPT-4o-mini fall between these extremes, with effect sizes closer to the human baseline than to the zero-shot LLMs.

To quantify this, we compare the absolute deviation of each paradigm's ATE from the Human ATE across the seven fully significant topics (Donald Trump, Healthcare, Republican Party, Joe Biden, Barack Obama, Hillary Clinton, Democratic Party). Fine-tuned GPT-4o-mini is the closest to Human, with a mean absolute deviation of $0.004$ across the seven topics. Fine-tuned RoBERTa is the next closest at $0.006$. GPT-4o-mini (baseline) deviates by $0.014$ on average, and Llama-3.3-70B by $0.021$. The same pattern holds in the mediation analysis: Human's TE is $+0.135$, Fine-tuned RoBERTa's is $+0.154$ (deviation $0.019$), Fine-tuned GPT-4o-mini's is $+0.158$ (deviation $0.023$), while GPT-4o-mini baseline ($+0.255$, deviation $0.120$) and Llama-3.3-70B ($+0.316$, deviation $0.181$) diverge substantially.

This pattern suggests that fine-tuning, whether via continued pre-training (Fine-tuned RoBERTa) or supervised instruction tuning (Fine-tuned GPT-4o-mini), helps calibrate model predictions toward human-scale effect sizes. The zero-shot LLMs systematically overestimate the magnitude of the sentiment-ideology relationship, producing ATEs that are $2$--$2.5\times$ the human baseline on topics like Donald Trump. Fine-tuning compresses these inflated effects toward the human range without altering their direction or significance. In this sense, fine-tuning serves as an alignment mechanism for causal effect magnitudes: it does not change \emph{which} topics are significant or \emph{which direction} effects point, but it modulates \emph{how large} the estimated effects are, bringing them closer to the scale that human annotators produce.

The practical implication is that researchers using LLM annotations for downstream causal analyses should be aware that zero-shot LLM labels will systematically inflate estimated effect sizes relative to human judgments. Fine-tuned models provide a closer approximation, with Fine-tuned GPT-4o-mini producing the tightest match to human effect ranges across the topics examined here.

\subsubsection{Classification Accuracy Does Not Predict Causal Fidelity}

A notable feature of the effect size comparison is the disconnect between classification performance and causal effect magnitudes. Fine-tuned GPT-4o-mini achieves the highest F1-Macro among the machine annotators (72.48, Table~\ref{tab:llm-causal-performance}), substantially outperforming Fine-tuned RoBERTa (62.19), a gap of over 10 F1 points. Yet the two models produce strikingly similar mediation effects on the Donald Trump pathway: Fine-tuned GPT-4o-mini estimates a Total Effect of $+0.158$ ($Q25 \to Q75$), while Fine-tuned RoBERTa estimates $+0.154$ (Table~\ref{tab:te-trump}). The NDE/NIE decompositions are similarly close ($0.141/0.017$ vs.\ $0.135/0.019$), and both models agree on the proportion mediated through Politics sentiment (${\sim}11$--$12\%$).

This convergence despite a large F1 gap reinforces the finding from Section~\ref{sec:effect-magnitudes} that the two estimators (multi-treatment ATE and mediation TE) capture different aspects of the sentiment--ideology relationship. If causal effect magnitudes tracked classification accuracy, Fine-tuned GPT-4o-mini should produce systematically larger or more precise effects than Fine-tuned RoBERTa. Instead, the near-identical mediation profiles suggest that above a certain threshold of classification competence, additional F1 gains reflect improvements in discrimination among ambiguous articles (particularly center-leaning content, where GPT-4o-mini's advantage is concentrated) rather than changes in how the model encodes the sentiment--ideology causal pathway. F1 measures whether the model assigns the correct label; the causal analysis measures which textual features drive that assignment. These are orthogonal evaluation axes, and high performance on one does not guarantee alignment on the other.

\subsection{Implications for LLM Annotation}

A growing line of work explores the potential of LLMs as replacements for human annotators \citep{horton2023large, ma-etal-2025-algorithmic, gao2025LLMscaution}. Our findings provide a nuanced picture of this substitution. On the positive side, all five annotation paradigms agree on the direction and significance of effects for the most important partisan topics: no paradigm produces a false positive or a directional reversal on any of the seven fully significant topics. This suggests that for detecting the presence and direction of causal relationships between topic sentiment and ideology, LLM annotations are reliable proxies for human labels.

However, the magnitude of estimated effects varies substantially across paradigms. If LLM annotations are used as silver-standard training data or as inputs to causal analyses, zero-shot labels will systematically overestimate effect sizes. To make this concrete: consider a researcher analysing whether Donald Trump sentiment causally drives ideology classification. Using Llama-3.3-70B labels, they estimate an ATE of $+0.1055$; using Human labels, the estimate is $+0.0427$, roughly $2.5\times$ smaller. Both are significant and directionally identical, but the magnitude difference could lead to different conclusions about the practical importance of the effect. Fine-tuned models mitigate this: Fine-tuned GPT-4o-mini ($+0.0448$) and Fine-tuned RoBERTa ($+0.0449$) both fall within $5\%$ of the human estimate on this topic.

The non-significant topics reveal a second concern. When effects are small and non-significant, annotators disagree not just on magnitude but on direction (Table~\ref{tab:multi-nonsig}). Russia, Politics, and Elections all show directional splits across paradigms. In these cases, the choice of annotation paradigm could determine whether a researcher concludes that a positive or negative relationship exists, even though neither conclusion would be statistically significant. This directional instability on non-polarising topics reinforces the need for multi-paradigm validation when LLM annotations are used in causal pipelines.

\subsection{Differences Across LLM Families}

The annotators exhibit different causal profiles despite all predicting ideology from the same raw article text, suggesting that architecture and training data differences shape which causal pathways models rely on during inference.

Fine-tuned RoBERTa, despite being fine-tuned on the same AllSidesV2 ideology labels as Fine-tuned GPT-4o-mini, produces effect sizes that are consistently among the closest to Human. Two non-exclusive mechanisms could explain this. First, the encoder-only architecture may be structurally less prone to encoding inflated sentiment-ideology associations than the generative GPT family. Second, the metalinguistic continued-pretraining objective (Chapter~\ref{sec:continued-pretraining}) may regularise the representation space in a way that resists the magnitude amplification observed in zero-shot generative models. We cannot disentangle the two without an ablation we did not run, but the practical implication stands: encoder-based fine-tuning with auxiliary objectives produces effect estimates that closely track human magnitudes.

GPT-4o-mini (baseline) and Fine-tuned GPT-4o-mini share the same underlying architecture, so the divergence between them is attributable entirely to fine-tuning. On the seven fully significant topics, fine-tuning reduces the mean absolute ATE from $0.044$ (baseline) to $0.036$ (fine-tuned), compressing effect sizes toward the human range ($0.033$). The directional agreement is perfect: both paradigms assign the same sign to every significant topic. The fine-tuning process therefore acts as a magnitude attenuator rather than a directional corrector, consistent with the interpretation that zero-shot LLMs encode the correct causal structure but at inflated scale.

Llama-3.3-70B consistently produces the largest ATEs across all significant topics, with a mean absolute ATE of $0.055$ on the seven fully significant topics, roughly $1.7\times$ the human baseline. Despite this magnitude inflation, Llama agrees directionally with all other paradigms on every significant topic and produces no false positives on non-significant topics. Its causal profile is therefore best characterised as a scaled-up version of the human signal rather than a qualitatively different one.

\subsection{Shortcut Learning and Fine-Tuning}
\label{sec:shortcut-learning}

The community-level results provide direct evidence that fine-tuning can address shortcut learning artifacts in LLM annotation. Community~1 (Elections and Campaigns) is the clearest illustration. Baseline GPT-4o-mini detects a significant positive ATE ($+0.034$, 95\% CI $[+0.019, +0.050]$), but Fine-tuned GPT-4o-mini does not ($+0.009$, $[-0.010, +0.026]$). Human annotations are also non-significant ($+0.016$, $[-0.002, +0.034]$). The baseline model appears to rely on surface-level lexical cues associated with electoral content that spuriously correlate with ideology in the zero-shot setting, producing a significant effect where none exists under human judgment. Fine-tuning on human ideology labels removes this artifact, bringing the Fine-tuned GPT estimate into alignment with the human baseline: the signal detected by GPT-4o-mini baseline is not a genuine ideological effect but a shortcut that supervised training corrects.

This finding aligns with recent work on shortcut learning in LLMs. \citet{yuan2024llmsovercomeshortcutlearning} demonstrate that shortcut reliance persists in LLMs even in few-shot settings, and \citet{sun-etal-2024-exploring} confirm that prompting alone is insufficient to eliminate spurious feature reliance, highlighting the importance of fine-tuning. Both studies identify chain-of-thought (COT) reasoning prompts as particularly effective at mitigating shortcuts by forcing models to explicitly reason through intermediate steps rather than relying on surface correlations.

The role of fine-tuning is not uniformly corrective, however. In Community~0 (White House and Congress), fine-tuning has the opposite effect: baseline GPT is non-significant ($-0.013$) while Fine-tuned GPT reaches significance ($-0.025$), moving the estimate closer to the human value ($-0.024$). Here, fine-tuning amplifies sensitivity to a genuine signal rather than removing a spurious one. The contrasting behaviour across communities is consistent with a dual role for fine-tuning: it suppresses shortcut-driven artifacts (Community~1) while strengthening genuine ideological signals (Community~0). Llama-3.3-70B's large significant effect in Community~1 ($+0.056$) further supports this interpretation: as a zero-shot model without task-specific fine-tuning, Llama may be subject to the same surface-level cue reliance as baseline GPT, but at a larger magnitude.

The Donald Trump mediation analysis (Table~\ref{tab:te-trump}) provides a subtler illustration of the same phenomenon. All four machine annotators produce significant Total Effects on the $Q25 \to Q75$ contrast: Fine-tuned RoBERTa ($+0.154$), GPT-4o-mini baseline ($+0.255$), Fine-tuned GPT-4o-mini ($+0.158$), and Llama-3.3-70B ($+0.316$). Human annotations, by contrast, narrowly miss significance ($+0.135$, 95\% CI $[-0.001, +0.270]$). The pattern suggests that all four machine paradigms overestimate the strength of the Trump-sentiment pathway relative to human judgment, and that fine-tuning reduces but does not fully eliminate this inflation. Fine-tuned RoBERTa and Fine-tuned GPT-4o-mini produce effects much closer to the human estimate than their zero-shot counterparts, yet both still cross the significance threshold that the human baseline does not. This residual gap indicates that fine-tuning attenuates shortcut-driven amplification of the sentiment signal without completely removing it.

Disentangling the sources of this residual inflation remains an open question. For Fine-tuned RoBERTa, the metalinguistic continued-pretraining objective and the encoder architecture are confounded: both could contribute to the model's sensitivity to sentiment cues. Running the same causal analysis with a baseline RoBERTa (fine-tuned on AllSidesV2 without continued pre-training) would isolate the contribution of the metalinguistic objective by holding architecture constant. Similarly, evaluating the POLITICS model \citep{liu2022politics}, which uses contrastive triplet-loss pre-training on political text without any sentiment-specific objective, would test whether domain-adapted encoder models in general exhibit this residual inflation or whether it is specific to the metalinguistic training signal. These ablations would clarify whether the near-significance gap between fine-tuned machine annotators and human judgments reflects an architectural tendency of transformer classifiers, a byproduct of sentiment-aware pre-training, or some combination of both.

Our OPRO analysis (Appendix~\ref{appendix:opro}) found that COT prompts slightly underperformed the baseline on F1 metrics (51.47 vs.\ 51.86). However, F1 measures classification accuracy, not causal fidelity. A model that reasons more carefully about ideology could produce different causal effect patterns even if its overall accuracy is comparable. Evaluating COT-generated labels through the causal framework of Chapter~\ref{sec:causal-inference} could determine whether explicit reasoning at inference time serves as an alternative to fine-tuning for aligning causal estimates with human judgments. We identify this as a direction for future work in Section~\ref{sec:future-directions}.

Among the significant communities, fine-tuned models consistently produce effect sizes closest to the human baseline. In Community~0, Fine-tuned GPT ($-0.025$) deviates from Human ($-0.024$) by only $0.001$, and Fine-tuned RoBERTa ($-0.023$) deviates by $0.001$. In Community~2 (Public Health), Fine-tuned GPT ($-0.031$) is the closest paradigm to Human ($-0.036$), with a deviation of $0.005$. Across both communities, the fine-tuned models outperform their zero-shot counterparts in alignment with human effect magnitudes, reinforcing the finding from Section~\ref{sec:causal-discussion} that fine-tuning serves as a magnitude attenuator that brings LLM-derived causal estimates closer to human scale.

\subsection{Mediation Analysis as Diagnostic Framework}

A central challenge in deploying LLMs as annotators is that in-distribution F1 alone does not expose why a model produces a given label, and the central claim of this chapter should be stated with that scope. F1 is not the only evaluation tool available: out-of-distribution robustness tests, cross-dataset evaluation, and adversarial probing can sometimes catch the same spurious feature reliance that causal probing catches, and we do not claim otherwise. Causal probing complements robustness benchmarks rather than replacing them, and is most useful in exactly the setting we examine here, where in-distribution accuracy is high but the underlying pathway from sentiment to ideology may diverge from the human one. \citet{moraffah2020causal} argue that causal interpretability, understanding decisions in terms of which inputs caused which outputs, is essential for building trustworthy ML systems, precisely because correlation-based evaluation leaves spurious feature reliance invisible.

External causal probing is therefore structurally necessary, not merely complementary to accuracy evaluation. \citet{jin2024largelanguagemodelsinfer} show that LLMs perform near-randomly on tasks requiring pure causal inference from correlational evidence, and \citet{joshi-etal-2024-llms} demonstrate that LLMs are prone to post hoc and ordering fallacies when reasoning about causation. \citet{fu2025correlationcausationanalyzingcausal} further show that standard LLMs rely on spurious correlations in their reasoning processes rather than genuine causal structure, a deficiency that RLVR training partially mitigates but does not eliminate. Taken together, these results imply that LLMs cannot be relied upon to self-report or self-correct the causal basis of their annotations; external causal analysis is required.

Our mediation framework provides one such approach. By holding the input text and causal structure fixed and varying only the annotation paradigm, we isolate differences that are attributable to the annotator rather than the data. This design is consistent with the evaluation principles articulated by \citet{cheng2022evaluation}, who emphasize the need for structured causal evaluation pipelines that go beyond conventional accuracy benchmarks. Applied to annotation auditing, mediation analysis can surface which textual features drive predictions through direct versus mediated pathways, offering a principled basis for trustworthiness assessment that F1 and agreement metrics cannot provide.
