\subsection{Causal Inference in NLP}
\label{sec:rw-causal-inference-nlp}

Causal inference methods have been applied to text analysis across a range of settings, and the literature is usefully organized by the role text plays in the causal graph. In \emph{text-as-treatment} settings, features extracted from text (sentiment, framing, topic) are treated as the intervention whose effect on a downstream outcome is estimated. In \emph{text-as-outcome} settings, the text itself is the response variable, and the question is how some non-textual treatment shifts language. In \emph{text-as-confounder} settings, text is used to adjust for unobserved structure that would otherwise bias an estimated effect. \citet{feder2022causalinferenceinnlp} provide a comprehensive survey across these settings; \citet{veitch2019textembeddingscausalinference} propose using text embeddings to adjust for confounding; \citet{keith-etal-2021-text} develop a framework for text as causal mediator, estimating direct and indirect effects of social group signals through language; and \citet{tierney-volfovsky-2021-sensitivity} apply causal mediation analysis to political polarization through text. The causal analysis in Chapter~\ref{sec:causal-inference} sits firmly in the text-as-treatment regime: text-derived sentiment and topic features play the role of treatment, and the human or model-assigned ideology label plays the role of outcome.

In parallel, causal methods have been applied to understand the behavior of language models themselves. \citet{vig2020genderbiasLLMs} use causal mediation analysis to localize and diagnose gender bias in language models, decomposing how gendered surface cues propagate through model components into biased outputs. \citet{stolfo2023mechanisticinterpretationarithmeticreasoning} apply mechanistic interpretation via causal mediation to arithmetic reasoning, and \citet{han2024surfacestructurecausalassessment} use causal probing to assess LLM comprehension beyond surface structure. The structural analogy to this thesis is direct: the gender-bias mediation work asks whether a spurious surface feature (gendered tokens) causally drives a downstream classification, while our analysis asks whether a spurious surface feature (article sentiment) causally drives a downstream ideology classification. In both cases, mediation is the diagnostic tool by which a hidden shortcut becomes measurable.

Our work in Chapter~\ref{sec:causal-inference} contributes to this literature by applying causal inference in a previously unexplored role: as an \emph{annotation-quality diagnostic} rather than as a model-internals interpretability tool. Instead of opening up a single model and asking which neurons mediate a behavior, we take entire annotation pipelines (human, fine-tuned encoders, zero-shot LLMs, fine-tuned LLMs) as black-box label generators and ask whether the causal structure linking sentiment, topic, and ideology is preserved across them. Divergences in that causal structure (not divergences in F1) are what reveal shortcut learning. To our knowledge, framing causal mediation as a downstream auditor of annotator trustworthiness, rather than as a probe of a fixed model's internals, is a novel use of these methods in NLP.
