\section{Computational Bias Detection}

The computational detection of political bias in news media has evolved from simple keyword-based approaches to sophisticated neural architectures. Early work in this domain focused primarily on lexical and syntactic features, often employing traditional machine learning methods such as Support Vector Machines and logistic regression \citep{recasens2013weasel}. These approaches, while interpretable, suffered from limited capacity to capture complex semantic relationships and contextual dependencies.

The advent of pre-trained language models has transformed the landscape of bias detection, but also surfaced a fundamental failure mode: source leakage. When fine-tuned naively on article-ideology pairs, transformer encoders learn to identify the publishing outlet from stylistic and lexical cues and route predictions through that outlet's known ideological label, rather than learning content-grounded ideology. The resulting models generalize poorly to unseen sources and inflate in-distribution F1 in ways that mask the lack of true ideological understanding. Contrastive triplet-loss objectives were introduced to address this failure mode by forcing ideologically similar articles from \emph{different} sources to cluster in embedding space, breaking the source--ideology shortcut. \citet{baly2020we} instantiated this approach by introducing BERT-based triplet-loss training for political ideology classification. \citet{liu2022politics} subsequently scaled the methodology with large-scale continued pre-training and same-story contrastive objectives, producing the POLITICS model.

Recent work by \citet{ronnback2025biasoutlets} has explored the integration of semantic, syntactic, and metalinguistic features for source-level bias detection, demonstrating the potential value of multi-modal feature representations. However, the systematic integration of such features into article-level classification systems remains underexplored.

At the article level, only AllSides provides ratings that do not always align with source-level bias. \citet{baly2020we} constructed the AllSides dataset and addressed source leakage through triplet-loss pre-training for BERT, using contrastive learning to structure embeddings by ideological similarity rather than source-identifying patterns. \citet{liu2022politics} scaled this approach with a larger pre-training dataset and refined the contrastive objectives to focus on same-story article comparisons across sources, producing the POLITICS model with current state-of-the-art results on AllSides. Our work extends this paradigm with metalinguistic auxiliary objectives.

\subsection{Metalinguistic Awareness in Political Text Analysis}

Metalinguistic awareness, i.e. the capacity to reflect upon and manipulate language as an object of thought, rather than simply using it for comprehension and production, is foundational to human reading comprehension and language learning \citep{tunmer1984development, nagy2007metalinguistic, Zipke10092007}. In the context of political text, metalinguistic features capture \textit{how} language is used to encode stance and ideology, as distinct from the propositional content of what is said. This distinction is central to our approach. The mechanism by which metalinguistic supervision is expected to improve ideology classification is representational rather than purely additive: by forcing the encoder to predict explicit metalinguistic targets (theme labels and tone scores) alongside the masked-language objective during continued pre-training, the model is regularized toward an internal representation space in which stylistic and topical structure are salient and linearly separable. This in turn aligns the encoder's internal geometry with the cues that human annotators report drawing on when making ideological judgments (emphasis, framing, and affective stance), so that the downstream classifier has access to features that mirror the human decision process rather than being forced to recover them implicitly from token statistics. We accordingly hypothesize that models trained to recognize metalinguistic patterns, specifically thematic emphasis and affective tone, will develop richer representations of political orientation.

In NLP, sentiment and thematic content have been studied as ideology signals across a range of settings. \citet{mullen2006preliminary} applied sentiment analysis to informal political discourse, finding that standard text classification methods are insufficient for capturing ideological orientation and motivating richer linguistic signals. \citet{smirnova2017ideology} used lexicon-based sentiment analysis (LIWC) to trace ideological shifts in New York Times coverage, demonstrating that tonal patterns correlate with editorial positioning. \citet{bhatia2018topic} showed that topic-specific sentiment polarities identify ideology in both congressional speeches and news articles, establishing a direct link between thematic framing and political orientation. More recently, \citet{bestvater2023sentiment} established that overall sentiment and target-directed stance diverge in political texts, underscoring the complexity of affective signals in this domain. At the sentence level, \citet{recasens2013weasel} showed that stylistic features can identify potentially biased language.

Building on this line of work, \citet{ronnback2025biasoutlets} demonstrated that semantic, syntactic, and metalinguistic features, including tone and topic, classify the political bias of news outlets at 76\% accuracy against a 45\% majority-class baseline.

These threads can be organized into three distinct claims that the present thesis builds on. First, sentiment, tonal patterns, and topic-specific stance \emph{correlate with} ideology in political text \citep{mullen2006preliminary, smirnova2017ideology, bhatia2018topic, bestvater2023sentiment}. Second, when aggregated to the source level, metalinguistic features are sufficient to \emph{classify outlets} along the political spectrum \citep{ronnback2025biasoutlets}. Third (and this is the open gap that this thesis fills), prior work has not systematically integrated these features into \emph{article-level} classifiers via auxiliary pre-training objectives. Our continued pre-training framework supplies exactly this missing step, attaching theme and tone heads to a contrastive backbone so that metalinguistic signal is injected into the encoder's representation space rather than being treated as a source-aggregated post-hoc statistic.
