\section{LLMs as Surrogates for Human Judgment}
\label{sec:rw-llm-surrogates}

A growing body of research explores whether large language models can serve as surrogates for human participants across social science tasks. \citet{horton2023large} frames LLMs as simulated economic agents whose behavior can be studied in experimental settings, while \citet{li2024frontiers} investigate the validity of LLMs for automated perceptual analysis in marketing research. In political science specifically, \citet{wu2023large} demonstrate that LLMs can estimate the latent political positions of politicians from legislative text, \citet{ibrahim2024analyzingpoliticalstancestwitter} show promise for political stance detection in social media, and \citet{strachan2024testing} test theory of mind capabilities in LLMs compared to human subjects. \citet{ma-etal-2025-algorithmic} further evaluate the \emph{algorithmic fidelity} of LLM-generated synthetic public opinions (that is, whether LLM-generated synthetic opinions reproduce the statistical distribution of real human opinions), finding that model outputs can approximate aggregate opinion distributions under constrained conditions. The potential of prompt engineering and few-shot learning approaches for political classification is particularly appealing, as these methods can leverage the extensive world knowledge encoded in LLMs to perform ideology prediction without task-specific training data, effectively substituting for costly human annotation. This thesis evaluates modern LLMs on political bias classification across both zero-shot and fine-tuning paradigms (Chapters~\ref{sec:baselines} and \ref{sec:llm-finetuning}), and then examines whether the \textit{causal structure} of their predictions, specifically, which textual features drive ideology judgments, mirrors that of human annotators (Chapter~\ref{sec:causal-inference}).

Despite their promise, several fundamental limitations constrain the reliability of LLMs as substitutes for human judgment. Transferability remains an open concern: \citet{gao2025LLMscaution} demonstrate that in economic games requiring strategic reasoning, LLMs consistently fail to replicate the human behavior distribution, with failure modes that vary inconsistently across models and input variations, suggesting that surrogate validity established in one domain does not straightforwardly transfer to another. Data contamination poses a particular risk for political classification tasks. Public ideology-labeled resources such as AllSides, Media Bias/Fact Check, and the AllSides-derived datasets used in prior bias research are exactly the kind of curated, widely linked corpora that are likely to appear in major LLM training pipelines. As a result, an LLM's apparent ability to classify article ideology may reflect memorized source--ideology pairings rather than learned content-based reasoning. This concern parallels the source leakage problem identified by \citet{baly2020we} in supervised classifiers, but is more difficult to diagnose in LLMs due to the opacity of their training data, and it is the specific reason that this thesis turns to causal probing in Chapter~\ref{sec:causal-inference} rather than relying on F1 alone to certify LLM annotators.

Shortcut learning further undermines the robustness of LLM predictions. \citet{yuan2024llmsovercomeshortcutlearning} show that LLMs remain susceptible to spurious correlations, relying on surface-level lexical patterns rather than deeper semantic understanding. In the political domain these shortcuts take concrete and predictable forms: salient named entities (Trump, Biden, Pelosi), ideologically marked keywords (\emph{socialism}, \emph{freedom}, \emph{woke}, \emph{patriot}), and even publication names that surface inside quoted material can act as high-precision proxies for the gold ideology label, allowing a model to score well in distribution without engaging the article's framing at all. and \citet{zhou2024explorespuriouscorrelationsconcept} demonstrate that such concept-level spurious correlations persist even in large-scale language models for text classification. In the political domain, this manifests as sensitivity to named entities, topic cues, or stylistic markers that correlate with ideology in training data but do not constitute genuine ideological signals. In media bias classification specifically, \citet{Wen_2023} showed that GPT-3.5 underperforms fine-tuned encoder-only models on directional ideology classification, despite remaining competitive on simpler binary stance and topic-classification tasks. This is one of the strongest existing pieces of evidence that headline LLM capability does not transfer to article-level ideology in particular, and it directly motivates the structure of our evaluation: rather than treating LLMs as a single monolithic baseline, we compare zero-shot and fine-tuned LLM variants against fine-tuned specialized encoders on the same articles, so that the specific gap Wen identified can be measured under controlled conditions. Our empirical comparison of fine-tuned transformers, zero-shot LLMs, and fine-tuned LLMs directly addresses this gap in the context of article-level political ideology classification.

Finally, low interpretability limits the ability to audit LLM-based surrogate annotations. Unlike human annotators, who can articulate their reasoning, or traditional classifiers with established feature importance methods, LLMs operate as opaque decision-makers whose internal processes resist straightforward inspection \citep{moraffah2020causal}. This opacity is especially problematic when LLM annotations are used as inputs to downstream causal analyses, where systematic biases in annotation can propagate into, or entirely reverse, estimated causal effects. The causal analysis in Chapter~\ref{sec:causal-inference} directly addresses this concern by comparing the causal structure of predictions across five annotation paradigms.
