\section{Background and Motivation}

In the current digital landscape, the proliferation of politically biased news content presents substantial challenges for democratic discourse and informed decision-making. The increasing polarization of media sources, combined with algorithmic content curation systems, has created online ecosystems where algorithmic amplification exposes participants to ideologically diverse content in ways that nonetheless drive political polarization \citep{flaxman2016filterbubblesechochambers,kitchens2020echochambersfilterbubbles,Garimella_Smith_Weiss_West_2021, cinelli2021echochambereffectsocialmedia}. As people are generally worse at identifying their own biases \citep{wang2020bias} than those of others, this phenomenon raises concerns about how individuals consume, interpret, and share news information.

Computational systems that surface ideological signal in news text can support media literacy, content curation, and journalistic accountability, which is what justifies treating bias detection as a research problem. The automatic detection and classification of political bias in textual content has accordingly emerged as a recent research area within natural language processing and computational social science. Such systems have practical applications in media literacy tools that help readers recognize ideological framing in the articles they encounter, recommendation systems that diversify exposure across the ideological spectrum, fact-checking pipelines that flag ideologically loaded framing for human review, and journalistic analysis workflows that audit coverage patterns across outlets. However, the computational modeling of political ideology remains a challenging problem due to the inherent subjectivity of bias perception, the complex relationship between explicit content and implicit framing, and the need to distinguish between source-level and article-level ideological markers.

Traditional approaches to bias detection have primarily focused on source-level classification, where entire news outlets or authors are categorized according to their general political orientation \citep{baly2020acl, darwish2020aaai, ronnback2025biasoutlets}. While this coarse-grained approach provides useful insights for media analysis, it fails to capture the significant variation in political perspective that can occur within individual articles from the same source. Article-level classification represents a more nuanced and practically valuable approach, enabling fine-grained analysis of ideological content and reducing the risk of overgeneralization based on source reputation alone.

Evidence from prior work motivates theme and tone as the specific auxiliary objectives we investigate in this thesis. Metalinguistic features, which include properties such as affective tone and thematic content that reflect \textit{how} language is used and not just what is said, have been shown to carry ideological signal at the source level (see Section~\ref{sec:relatedwork} for a full review). This is consistent with cognitive science research establishing that metalinguistic awareness, the capacity to reflect on and manipulate the structural features of language as an object of thought, is foundational to human reading comprehension \citep{tunmer1984development, nagy2007metalinguistic, Zipke10092007}. These results suggest that metalinguistic signals mirror what human annotators draw on when assessing political orientation, and motivate our approach of incorporating theme prediction and tone regression as auxiliary heads during continued pre-training.

One might object that source-level evidence does not automatically transfer to the article level, since articles within a single outlet can vary widely in ideology and tone. We argue that the transfer holds precisely because of how the auxiliary supervision is applied: by attaching explicit per-article theme and tone targets during continued pre-training, the encoder is forced to develop fine-grained representations of each article's metalinguistic profile rather than relying on outlet-level memorization of source--ideology associations. The auxiliary objectives push the model to distinguish within-source ideological variation, which is what an article-level classifier requires.

We note that the literature uses varying terminology: \citet{ronnback2025biasoutlets}, \citet{baly2020we}, and crowd-sourced resources such as AllSides and MB/FC refer to political ``bias'', while \citet{liu2022politics} use political ``ideology''. Throughout this thesis, when discussing specific works or resources we adopt their own terminology; when describing our own contributions we prefer ``ideology''.

While large language models (LLMs) are increasingly used as proxies for human judgment in social science tasks \citep{wu2023large, horton2023large, li2024frontiers, ma-etal-2025-algorithmic}, \citet{gao2025LLMscaution} caution against treating LLMs as human surrogates, demonstrating that LLM behavior distributions fail to replicate those of humans across multiple domains. This thesis provides an empirical comparison of our continued pre-training approach against LLMs in both zero-shot and fine-tuned settings to characterize this gap concretely, and further examines whether the causal structure of LLM predictions mirrors that of human annotators through the causal analysis framework presented in Chapter~\ref{sec:causal-inference}.

News outlets across the political spectrum cover the same events differently: not just in which topics they emphasize, but in the sentiment they express toward those topics. A growing body of work has examined this relationship between sentiment, topic coverage, and ideological stance \citep{smirnova2017ideology, bhatia2018topic, bestvater2023sentiment}. \citet{smirnova2017ideology} found that \textit{New York Times} coverage of Russia and Islam shifted toward more negative sentiment following the Crimea annexation and 9/11, respectively. \citet{bestvater2023sentiment} caution that tone alone is an imperfect predictor of political stance, but demonstrate that in specific contexts the correlation between sentiment and ideological alignment is substantial. Our own results in Chapter~\ref{sec:continued-pretraining} provide further support: incorporating thematic and tonal signals during pre-training yields consistent improvements in ideology classification.

Table~\ref{fig:khalil} illustrates the phenomenon concretely. AP and Fox News both covered activist Mahmoud Khalil's arrest using the same AllSides topic tags, yet their language diverges sharply. AP describes ``demonstrators'' who ``filled the lobby'' to ``denounce'' an ``immigration arrest''; Fox describes ``protesters'' who ``occupied'' the lobby, labeling Khalil an ``anti-Israel activist'' who was ``detained.'' Even the headlines encode stance: AP uses ``flood'' and explains the protesters' objective; Fox uses ``occupy'' without contextualizing their purpose. Sentiment toward the same topics (Pro-Palestine Protests, Israel, Deportations) can serve as an indicator of underlying ideological framing.

\begin{table}[h!]
    \centering
    \begin{tabular}{p{2cm}|p{4cm} | p{4cm}}
\toprule
 &  Associated Press  & Fox News \\
 \midrule
topic tags &  Pro-Palestine Protests, Donald Trump, Israel, Deportations &  Pro-Palestine Protests, Donald Trump, Israel, Deportations \\
\midrule
title & Jewish protesters \textcolor{blue}{flood} Trump Tower's lobby to demand the Columbia University activist's release & Protesters supporting Mahmoud Khalil \textcolor{blue}{occupy} Trump Tower lobby \\
\midrule
summary &

\textcolor{blue}{Demonstrators} from a Jewish group filled the lobby of Trump Tower on Thursday to denounce the \textcolor{blue}{immigration arrest} of Mahmoud Khalil, a \textcolor{blue}{pro-Palestinian activist} who helped lead protests against Israel at Columbia University.

The \textcolor{blue}{Jewish Voice for Peace protesters}, who carried banners and wore red shirts reading "Jews say stop arming Israel," chanted "Bring Mahmoud home now!"
 &

About 150 \textcolor{blue}{protesters supporting Mahmoud Khalil} occupied the lobby area of Trump Tower on Manhattan's Fifth Avenue on Thursday, calling for the release of the \textcolor{blue}{anti-Israel activist} who was \textcolor{blue}{detained} over the weekend.

The protesters, many dressed in red, held signs reading "Free Mahmoud, Free Palestine" and "Fight Nazis Not Students." The protesters were also chanting the common "Free, free Palestine" slogan. \\
\bottomrule
\end{tabular}
\caption{AllSides reporting on the Khalil arrest. Words indicating sentiment or stance are highlighted in blue.}
\label{fig:khalil}
\end{table}

Despite this suggestive correlation, a critical gap remains: we do not know whether topic-specific sentiment \emph{causes} ideology classification or merely correlates with it. This distinction matters for two reasons. First, a causal analysis can identify which topics carry the strongest ideological signal and how those patterns shift over time. Second, applying causal methods to both human annotations and automated classifiers allows us to detect whether those systems capture genuine ideological signals or spurious associations.

\paragraph{Why F1 cannot detect this on its own.} It is worth being explicit about why standard accuracy metrics are blind to the question we are about to ask. F1, accuracy, and per-class precision/recall all measure the input--output mapping of a classifier: given an article, does the predicted label match the gold label? They evaluate what the model produces, not which features the model uses to produce it. In Table~\ref{fig:khalil}, AP and Fox News cover the Khalil arrest with identical topic tags but divergent framing, and two classifiers could agree on the ideology labels for both articles while relying on completely different internal pathways. One classifier might route its predictions through a human-aligned chain of framing, topical emphasis, and stance, recognizing the same linguistic cues a human annotator would weigh. The other might route its predictions through a spurious sentiment-only shortcut that happens to correlate with ideology in the training distribution, latching on to affective polarity without engaging the framing structure at all. In-distribution F1 cannot tell these apart, because both pathways produce the same labels on the same articles. A model can therefore achieve a high F1 by pushing predictions through a spurious sentiment pathway whose causal structure diverges sharply from the human pathway, and the classification scoreboard will register success either way. Causal probing addresses this gap: rather than asking whether the labels match, it asks whether the causal pathway from sentiment and topic to ideology matches the human pathway, and surfaces shortcut learning that F1 by construction cannot.
