\section{Cross-Paradigm Synthesis}
\label{sec:cross-paradigm-synthesis}

Having presented community-level and topic-level results separately, we now synthesize findings across both analysis levels and all five annotation paradigms.

\subsection{Agreement Patterns Across Analysis Levels}

At the \textbf{community level}, 14 cells reach significance across four of nine qualifying communities. Community~2 (Public Health) achieves full cross-paradigm replication: all five paradigms produce significant negative ATEs (range $-0.027$ to $-0.045$), matching the level of agreement seen among the partisan topics at the topic level. Community~0 (White House and Congress) produces three significant negative ATEs (Human, Fine-tuned RoBERTa, Fine-tuned GPT), Community~1 (Elections and Campaigns) produces two significant positive ATEs (baseline GPT, Llama), and Community~5 (Environment and Fiscal Policy) produces a single significant negative ATE (baseline GPT). Directional agreement is high: all five paradigms agree on the sign of the effect in each of the four significant communities.

At the \textbf{topic level}, the topic level analysis (Table~\ref{tab:multi-continuous}) shows thatseven topics produce significant effects that replicate across all or nearly all annotation paradigms: Donald Trump, Hillary Clinton, Democratic Party, Barack Obama, Republican Party, Healthcare, and Joe Biden. Community~2 (Public Health) matches this pattern at the community level with full cross-paradigm replication, but the remaining significant communities show more paradigm-specific patterns. The Donald Trump mediation analysis produces significant Total Effects under four of five paradigms (all except Human, which narrowly misses), with all five agreeing on a positive direction.

The key difference between the two analysis levels is the role of topic specificity. At the community level, each treatment aggregates sentiment across many related topics, diluting any single partisan signal. At the topic level, explicitly partisan topics (named political figures, parties, and policy domains like Healthcare) carry effects that are both larger and more replicable than any community-level finding. The non-partisan topics that dominate by article count (Elections, Politics, Presidential Elections) produce no significant effects at the topic level, mirroring the null results that characterise most communities.

\subsection{Interpreting Effect Magnitudes}
\label{sec:effect-magnitudes}

Across both analysis levels and all five annotation paradigms, the estimated ATEs are uniformly small relative to the $[0, 2]$ ideology scale. At the community level, significant ATEs range from $0.019$ (Community~5, baseline GPT) to $0.056$ (Community~1, Llama), while at the topic level the largest significant ATE is $0.106$ (Donald Trump, Llama). Even the most robust finding, the Donald Trump multi-treatment ATE under Human annotations, amounts to only $+0.043$ on a two-point scale. These magnitudes suggest that topic-specific sentiment functions as a supplementary signal in ideology classification rather than a primary driver. The bulk of ideology prediction is carried by other features of article text that are absorbed into the confounders $W$; sentiment contributes a detectable but modest incremental effect, concentrated in explicitly partisan topics.

A striking feature of the results is the apparent discrepancy between the multi-treatment ATE and the mediation Total Effect for the same topic. Under Human annotations, the Donald Trump ATE from the multi-treatment model is $+0.043$ (Table~\ref{tab:multi-continuous}), while the mediation Total Effect is $+0.135$ (Table~\ref{tab:te-trump}), roughly three times larger. This gap illustrates how the two estimators answer fundamentally different causal questions and operate on different populations.

The multi-treatment ATE is estimated by a single \texttt{LinearDML} model that fits all 20 topic sentiments simultaneously as co-treatments. The Donald Trump coefficient captures only the variation in Donald Trump sentiment that is orthogonal to the remaining 19 topic sentiments after partialling them out. The model uses the full dataset, imputing zero sentiment for articles where a topic is absent and letting the topic-presence indicators in $W$ absorb that structural variation. Because the majority of articles in the corpus do not mention Donald Trump, much of the residualised treatment signal is concentrated near zero, diluting the estimated marginal effect.

The mediation Total Effect, by contrast, treats Donald Trump sentiment as the sole treatment and Politics sentiment as the mediator. By definition, the Total Effect does not control for the mediator: it measures the full causal flow from treatment to outcome, including whatever portion runs through the Politics channel. Moreover, the mediation pipeline restricts the sample to the $N = \num{1424}$ articles where both Donald Trump and Politics are tagged as present, a substantially smaller and more homogeneous population than the full dataset used by the multi-treatment model. Within this subpopulation, every observation has non-trivial Donald Trump sentiment variation, so the treatment signal is never diluted by structurally inert observations.

The multi-treatment ATE therefore answers: \emph{what is the effect of Donald Trump sentiment holding all other topic sentiments fixed?} The mediation TE answers: \emph{what is the total effect of Donald Trump sentiment, allowing it to also operate through Politics sentiment, among articles that discuss both topics?} The former is smaller precisely because it nets out shared variation with co-treatments and estimates the effect across a population where the treatment is mostly inert. The latter is larger because it preserves indirect pathways and conditions on a subpopulation where sentiment variation is meaningful in every observation.

The two estimators serve complementary roles. In the general news landscape, the multi-treatment ATE reflects the marginal predictive value of any single topic's sentiment across the full corpus. If one were to select two articles at random from AllSides and attempt to predict the difference in ideology labels, the ATE provides the best estimate of how much knowing each article's Donald Trump sentiment would improve that prediction, relative to all other sentiment signals competing for explanatory power. Because the prior probability of any given topic being present is low for most topics, this marginal contribution is necessarily small.

The mediation decomposition, estimated on the subpopulation where Donald Trump and Politics co-occur, addresses a different question. If one already knows that both articles discuss Donald Trump, one is conditioning on a population where sentiment variation is real and substantively meaningful in every observation. The Total Effect captures how much that sentiment matters in this narrower context, and the NDE/NIE decomposition reveals how much of the effect operates directly versus through the Politics channel. This decomposition is more informative for reasoning about the \emph{mechanism} by which sentiment shapes ideology classification, but only within articles where the topic is present.

The two quantities target genuinely different causal estimands on different populations. The ATE is the correct answer for assessing the general importance of a topic's sentiment across the news landscape, while the mediation decomposition is the correct answer for understanding mechanisms conditional on topic presence. Both are necessary for a complete picture: the small ATEs establish that sentiment is generally a supplementary rather than dominant signal, while the larger mediation effects reveal that within relevant subpopulations, the causal pathways from sentiment to ideology classification become substantively meaningful and amenable to mechanistic decomposition. Notably, the magnitude ordering across paradigms is not uniform: zero-shot LLMs consistently produce the largest effects while human annotations produce the smallest. We defer the analysis of paradigm-specific effect magnitudes, the relationship between classification accuracy and causal fidelity, and the implications for machine annotator reliability to Section~\ref{sec:causal-discussion}.
