\section{Topic-Level Results}
\label{sec:topic-level-results}

Having established that community-level aggregation reveals paradigm-specific effects (Section~\ref{sec:community-level-results}), we now present the finer-grained topic-level analysis for completeness. These results are available for five annotation paradigms: Human, Fine-tuned RoBERTa, GPT-4o-mini (baseline), Fine-tuned GPT-4o-mini, and Llama-3.3-70B-versatile.

\subsection{Mediation Analysis Results}
\label{sec:mediation-results}

We evaluate \textbf{Donald Trump sentiment as treatment} with Politics sentiment as mediator ($N = 233$ articles); the relatively small $N$ reflects the sparsity of Trump-tagged articles in the corpus.

Table~\ref{tab:te-trump} reports Total Effect estimates for both percentile contrasts across the five annotation paradigms. The treatment window is identical across all four datasets ($Q25 = -0.7$, $Q75 = +0.1$; $P10 = -1.1$, $P90 = +0.3$), so differences across paradigms arise entirely from differences in ideology labels $Y$.

\begin{table}[ht]
  \centering
  \sisetup{table-format=+1.3}
  \begin{tabular}{llS[table-format=+1.3]S[table-format=+1.3]S[table-format=+1.3]S[table-format=+1.3]S[table-format=+1.3]}
    \toprule
    \textbf{Paradigm} & \textbf{Contrast} &
      {\textbf{TE}} &
      {\textbf{CI lower}} & {\textbf{CI upper}} &
      {\textbf{NDE}} & {\textbf{NIE}} \\
    \midrule
    Human
      & $Q25 \to Q75$  & -0.254 & -0.934 & +0.426 & -0.160 & -0.094 \\
    Human
      & $P10 \to P90$  & -0.438 & -1.612 & +0.735 & -0.263 & -0.175 \\
    \addlinespace
    Fine-tuned RoBERTa
      & $Q25 \to Q75$  & -0.122 & -0.825 & +0.581 & -0.043 & -0.079 \\
    Fine-tuned RoBERTa
      & $P10 \to P90$  & -0.211 & -1.424 & +1.002 & -0.098 & -0.112 \\
    \addlinespace
    GPT-4o-mini (baseline)
      & $Q25 \to Q75$  & -0.034 & -0.208 & +0.141 & -0.113 & +0.079 \\
    GPT-4o-mini (baseline)
      & $P10 \to P90$  & -0.058 & -0.359 & +0.243 & -0.214 & +0.156 \\
    \addlinespace
    Fine-tuned GPT-4o-mini
      & $Q25 \to Q75$  & +0.087 & -0.522 & +0.695 & +0.043 & +0.044 \\
    Fine-tuned GPT-4o-mini
      & $P10 \to P90$  & +0.150 & -0.900 & +1.199 & +0.122 & +0.027 \\
    \addlinespace
    Llama-3.3-70B
      & $Q25 \to Q75$  & -0.045 & -0.233 & +0.143 & -0.022 & -0.023 \\
    Llama-3.3-70B
      & $P10 \to P90$  & -0.077 & -0.402 & +0.248 & -0.027 & -0.050 \\
    \bottomrule
  \end{tabular}
  \caption{Total Effect (TE) of Donald Trump sentiment on ideology label for
    two percentile contrasts. $N = 233$ articles. No estimate is statistically
    significant ($p < 0.05$); all 95\% CIs include zero.}
  \label{tab:te-trump}
\end{table}

The most striking finding is a \textbf{sign reversal} in the Total Effect: Human, Fine-tuned RoBERTa, and GPT-4o-mini (baseline) estimates are negative (more positive Trump sentiment $\to$ ideology shifts toward Left), while Fine-tuned GPT-4o-mini estimates are positive (more positive Trump sentiment $\to$ ideology shifts toward Right). Since the treatment window and article set are identical, this reversal reflects entirely how each labelling system classifies the same 233 articles. No estimate reaches statistical significance under any paradigm; all 95\% confidence intervals span zero. GPT-4o-mini (baseline) is notably more compact than the other negative paradigms: its TE ($-0.034$) and narrow CI ($[-0.208, +0.141]$) contrast with the wide intervals of Human and Fine-tuned RoBERTa, consistent with the baseline model's centrist prediction tendency compressing effect magnitudes. Its NDE ($-0.113$) is the largest in absolute value, while the positive NIE ($+0.079$) partially offsets it, a decomposition pattern distinct from the other paradigms.

The NDE accounts for the majority of the TE in all runs ($|\text{NDE}| > |\text{NIE}|$ in every case), indicating that most of the estimated effect does not route through the Politics sentiment mediator.

% ..............................................................................
\subsubsection{Conditional Average Treatment Effects by Outlet Ideology}
\label{sec:cate-results}
% ..............................................................................

To test whether the sentiment--ideology relationship differs by outlet type, we estimate CATEs using a Causal Forest DML~\citep{wager2018forests} and group results by outlet ideology (Left, Centre, Right). Heterogeneity here would indicate that the effect of sentiment on classification is moderated by the outlet's baseline ideology.

Table~\ref{tab:cate-trump} reports mean CATEs within each ideology group.

\begin{table}[ht]
  \centering
  \sisetup{table-format=+1.3}
  \begin{tabular}{l
      S[table-format=+1.3]@{\,}l
      S[table-format=+1.3]@{\,}l
      S[table-format=+1.3]@{\,}l}
    \toprule
    \textbf{Paradigm}
      & \multicolumn{2}{c}{\textbf{Left ($N$)}}
      & \multicolumn{2}{c}{\textbf{Centre ($N$)}}
      & \multicolumn{2}{c}{\textbf{Right ($N$)}} \\
    \midrule
    Human
      & -0.213 & {(80)}  & -0.211 & {(51)}  & -0.218 & {(102)} \\
    Fine-tuned RoBERTa
      & -0.141 & {(83)}  & -0.136 & {(49)}  & -0.139 & {(101)} \\
    GPT-4o-mini (baseline)
      & -0.058 & {(72)}  & -0.063 & {(121)} & -0.066 & {(40)}  \\
    Fine-tuned GPT-4o-mini
      & +0.056 & {(106)} & +0.069 & {(49)}  & +0.054 & {(78)}  \\
    Llama-3.3-70B
      & -0.073 & {(131)} & -0.069 & {(55)}  & -0.073 & {(47)}  \\
    \bottomrule
  \end{tabular}
  \caption{Mean CATE by outlet ideology group for Donald Trump sentiment treatment.
    The sign of the CATE reverses between Fine-tuned GPT-4o-mini and the other two paradigms.
    No significant between-group heterogeneity is detected within any single paradigm.}
  \label{tab:cate-trump}
\end{table}

Three observations stand out. There is \textbf{no ideology-based heterogeneity within any paradigm}: for both treatments, the CATE is nearly uniform across Left, Centre, and Right outlet groups, and the causal forest finds no evidence that the sentiment--ideology effect differs by outlet type. However, there is \textbf{large between-paradigm divergence}: the Fine-tuned GPT-4o-mini Politics CATE ($\approx +0.18$) is roughly three times the Fine-tuned RoBERTa CATE ($\approx +0.05$) and nine times the Human CATE ($\approx +0.002$), with GPT-4o-mini (baseline) falling between RoBERTa and Fine-tuned GPT ($\approx +0.07$), forming a clear magnitude gradient that is a property of the labelling models, not the articles. The \textbf{Trump CATE sign flip} is also notable: Human, Fine-tuned RoBERTa, and GPT-4o-mini (baseline) CATEs for Trump are uniformly negative, while Fine-tuned GPT-4o-mini CATEs are uniformly positive, mirroring the TE sign flip in Table~\ref{tab:te-trump}. GPT-4o-mini (baseline) produces the smallest negative CATEs ($\approx -0.06$), intermediate between RoBERTa ($\approx -0.14$) and the fine-tuned model's positive values, and its $N$ distribution is heavily centrist (121 of 233 classified as Centre), consistent with its known centrist prediction tendency.

% ------------------------------------------------------------------------------
\subsection{Multi-Treatment DML Results}
\label{sec:multi-treatment-results}
% ------------------------------------------------------------------------------

The multi-treatment sweep fits a single \texttt{LinearDML} model per dataset with all $k = 11$ qualifying topic sentiments as simultaneous treatments, controlling for topic-presence confounders via PCA-compressed $W$. ATEs are estimated on the full corpus ($N = \num{11516}$), providing substantially higher statistical power than the mediation subsample.

\subsubsection{Continuous Treatment Mode}
\label{sec:continuous-mode}

Table~\ref{tab:multi-continuous} reports the ATE of a one-standard-deviation increase in each topic's sentiment score, holding all other topic sentiments constant.

\begin{table}[ht]
  \centering
  \sisetup{table-format=+1.4}
  \setlength{\tabcolsep}{4pt}
  \small
  \begin{tabular}{l c
      S[table-format=+1.4]
      S[table-format=+1.4]
      S[table-format=+1.4]
      S[table-format=+1.4]
      S[table-format=+1.4]}
    \toprule
    \textbf{Topic} & {\textbf{N}}
      & {\textbf{Human}}
      & {\textbf{FT RoBERTa}}
      & {\textbf{GPT (base.)}}
      & {\textbf{FT GPT}}
      & {\textbf{Llama-3.3}} \\
    \midrule
    Hillary Clinton & 51 & +0.0051 & \cellcolor{blue!25}+0.0220 & -0.0064 & -0.0004 & -0.0034 \\
    Media Bias & 115 & +0.0024 & +0.0203 & -0.0130 & -0.0191 & +0.0057 \\
    FBI & 72 & +0.0182 & -0.0052 & -0.0006 & +0.0115 & -0.0077 \\
    Donald Trump & 233 & -0.0089 & -0.0025 & -0.0112 & +0.0011 & -0.0087 \\
    Media Watch & 58 & -0.0032 & -0.0079 & +0.0093 & +0.0078 & -0.0104 \\
    Russia Probe & 57 & +0.0011 & +0.0045 & -0.0002 & -0.0210 & +0.0100 \\
    Race and Racism & 104 & -0.0013 & +0.0040 & -0.0041 & +0.0050 & +0.0067 \\
    Defense and Security & 76 & +0.0017 & +0.0034 & -0.0076 & +0.0112 & +0.0068 \\
    LGBTQ Issues & 73 & +0.0027 & +0.0021 & -0.0045 & -0.0144 & -0.0063 \\
    Politics & 594 & -0.0035 & -0.0023 & +0.0098 & \cellcolor{blue!25}+0.0217 & +0.0017 \\
    Economy and Jobs & 56 & -0.0001 & +0.0019 & -0.0096 & -0.0095 & +0.0083 \\
    \bottomrule
  \end{tabular}
  \caption{Multi-treatment DML: ATE per topic under continuous encoding (one-SD increase in sentiment score). Cells shaded in blue denote statistically significant results (95\% bootstrap CI excludes zero; $B = \num{2000}$). The full table with 95\% confidence intervals for every estimate is reported in Appendix~\ref{sec:appendix-community-mediation}, Table~\ref{tab:multi-continuous-full}.}
  \label{tab:multi-continuous}
\end{table}

\subsubsection{The Hillary Clinton Signal}
\label{sec:hillary-signal}

Each paradigm identifies a different significant topic. Under Fine-tuned RoBERTa, the Hillary Clinton estimate narrowly excludes zero ($\widehat{\text{ATE}} = +0.022$, 95\% CI $[+0.001, +0.036]$), the only significant finding for this paradigm across all topic-level analyses. Under Fine-tuned GPT-4o-mini, only Politics reaches significance ($\widehat{\text{ATE}} = +0.022$, 95\% CI $[+0.003, +0.034]$). Neither finding appears in the Human or GPT-4o-mini (baseline) paradigms; both produce no significant topics. No topic is significant across more than one paradigm simultaneously. Outside of this isolated Hillary Clinton result, Fine-tuned RoBERTa produces no significant effects at the topic level, aligning with Human annotations throughout the mediation and CATE analyses.

The Hillary Clinton direction under Fine-tuned RoBERTa is counterintuitive: \emph{more positive} Clinton sentiment is associated with \emph{more right-leaning} ideology classification. A plausible explanation is selection bias: the 51 Clinton-tagged articles are disproportionately from 2016--2017, where outlets discuss her in a political-foil context (e.g., framing ``deep state'' criticism via Clinton references). That the effect is near zero under Fine-tuned GPT-4o-mini ($-0.0004$), slightly negative under GPT-4o-mini baseline ($-0.006$), and non-significant under Human labels strongly suggests the Fine-tuned RoBERTa signal is driven by labelling idiosyncrasies rather than a genuine causal relationship.

% ------------------------------------------------------------------------------
\subsection{Topic-Level Cross-Paradigm Comparison}
\label{sec:cross-paradigm}
% ------------------------------------------------------------------------------

\subsubsection{Agreement Across Paradigms}

Table~\ref{tab:cross-paradigm-summary} consolidates the directional agreement and significance of key causal estimates across all completed paradigm runs.

\begin{table}[ht]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{llcccccccccc}
    \toprule
    \textbf{Analysis} & \textbf{Topic} &
      \multicolumn{2}{c}{\textbf{Human}} &
      \multicolumn{2}{c}{\textbf{FT RoBERTa}} &
      \multicolumn{2}{c}{\textbf{GPT (baseline)}} &
      \multicolumn{2}{c}{\textbf{FT GPT-4o-mini}} &
      \multicolumn{2}{c}{\textbf{Llama-3.3-70B}} \\
    \cmidrule(lr){3-4}\cmidrule(lr){5-6}\cmidrule(lr){7-8}\cmidrule(lr){9-10}\cmidrule(lr){11-12}
    & & Dir. & Sig. & Dir. & Sig. & Dir. & Sig. & Dir. & Sig. & Dir. & Sig. \\
    \midrule
    Mediation (DT trt.)  & Trump $\to$ ideology
      & $-$ & No & $-$ & No & $-$ & No & $+$ & No & $-$ & No \\
    Multi-trt.  & Hillary Clinton
      & $+$ & No & $+$ & \textbf{Yes} & $-$ & No & $\approx 0$ & No & $-$ & No \\
    Multi-trt.  & Donald Trump
      & $-$ & No & $-$ & No & $-$ & No & $+$ & No & $-$ & No \\
    Multi-trt.  & Politics
      & $-$ & No & $-$ & No & $+$ & No & $+$ & \textbf{Yes} & $+$ & No \\
    \bottomrule
  \end{tabular}%
  }
  \caption{Summary of cross-paradigm agreement for key ATE estimates.
    ``Dir.'' = direction of ATE (positive/negative).
    ``Sig.'' = statistically significant at $\alpha = 0.05$.
    ``$\sim 0$'' = ATE indistinguishable from zero.}
  \label{tab:cross-paradigm-summary}
\end{table}

\subsubsection{Divergence Analysis}

Three patterns stand out. First, the Total Effect of Trump sentiment on ideology \textbf{reverses sign} between paradigms: negative under Human, Fine-tuned RoBERTa, and GPT-4o-mini (baseline), but positive under Fine-tuned GPT-4o-mini. Since the treatment window and article subset are identical across all four runs, this reversal reflects entirely how each model assigns ideology labels to the same 233 articles; annotation methodology alone can reverse an estimated causal direction. Notably, GPT-4o-mini (baseline) agrees with the human direction, suggesting the sign reversal is introduced specifically by the fine-tuning process.

Second, each fine-tuned model finds a \textbf{different significant topic} in the multi-treatment analysis, and no finding replicates in the human gold standard or baseline GPT. Fine-tuned RoBERTa finds Hillary Clinton significant; Fine-tuned GPT-4o-mini finds Politics significant; Human and GPT-4o-mini (baseline) find none. This pattern (one significant result per fine-tuned paradigm, pointing to different topics) is a strong indicator that the significant signals are annotation-model artifacts rather than robust causal findings.

% ------------------------------------------------------------------------------
\subsection{Topic-Level Summary}
\label{sec:causal-summary-topic}
% ------------------------------------------------------------------------------

The summary statement ``no topic produces a statistically significant effect that is consistent across annotation paradigms'' actually conflates two distinct empirical failures, and they are worth separating. The first is a within-paradigm null: \emph{within} most individual paradigms, no topic clears the significance threshold at all (Human, GPT-4o-mini baseline, and Llama-3.3-70B return no significant topic-level effects under any analysis). The second is a cross-paradigm disagreement: the few topics that \emph{do} reach significance in some paradigm fail to replicate in any other paradigm, and in some cases reverse direction (Fine-tuned RoBERTa's isolated Hillary Clinton result in the multi-treatment analysis is not recovered by any other paradigm; Fine-tuned GPT-4o-mini's positive Trump effect is the opposite sign of every other paradigm). The first failure is about statistical power and signal magnitude at the topic level; the second is about paradigm-dependence and is the structurally interesting finding. Across the topic-level analyses, then, no topic sentiment produces a statistically significant effect that is consistent across all annotation paradigms. Human, GPT-4o-mini (baseline), and Llama-3.3-70B annotations produce no significant effects across any analysis. Fine-tuned RoBERTa's only significant finding is an isolated Hillary Clinton result in the multi-treatment analysis that does not replicate under any other paradigm. The Donald Trump treatment direction is annotation-model dependent: negative under Human, Fine-tuned RoBERTa, GPT-4o-mini (baseline), and Llama-3.3-70B, but positive under Fine-tuned GPT-4o-mini. These results suggest that topic-level sentiment, as measured by current annotation models, is not a strong independent driver of ideology classification once topic presence is controlled for.

The primary value of the topic-level analysis lies in demonstrating that identical causal analyses applied to labels from different annotation systems can produce qualitatively different conclusions. As shown in the preceding community-level analysis (Section~\ref{sec:community-level-results}), aggregation across related topics does reveal more robust patterns, but these patterns remain paradigm-specific rather than universal.
