\chapter{Continued Pre-training Complete Results}
\label{sec:appendix-full-experiments}

Table~\ref{tab:pretraining-combined-full} presents the complete results for continued pre-training experiments across different triplet-loss configurations ($T_{id}(1)$, $T_{id}(16)$, $T_{id}(32)$) combined with theme ($T_h$) and tone ($T_o$) information, evaluated after epoch 1 and epoch 2. Results are shown after fine-tuning on the AllSidesV2 Media split.

\paragraph{How to read this table.} Three patterns are worth looking for as you scan the cells. (a) Cells highlighted in blue mark configurations where epoch~1 outperformed epoch~2, which is an early-stopping artifact rather than a normal training trajectory and is the primary anomaly we discuss below. (b) Center-class F1 stays weak across nearly all configurations, regardless of triplet sample size or auxiliary objective combination, which is the empirical fingerprint of the task-inherent center difficulty also seen in the human baseline (Section~\ref{sec:human-baseline}). (c) Configurations in which adding auxiliary heads \emph{degrades} the F1-Macro relative to the contrastive-only baseline are the empirical signature of the gradient-imbalance failure mode discussed in Chapter~\ref{sec:continued-pretraining}: when GDELT supervision is sparse, the auxiliary loss mostly contributes noise rather than signal at a fixed mixing weight.

\paragraph{The epoch-1 anomaly.} Several ``best'' cells in Table~\ref{tab:pretraining-combined-full} are reached at epoch~1 rather than epoch~2. This is one of the more interesting empirical patterns in Chapter~\ref{sec:continued-pretraining} and warrants its own paragraph rather than a single ``unexpected'' label. The most plausible mechanism is that, with very dense triplet supervision (large per-batch sample), the primary contrastive objective effectively converges within a single epoch, and a second epoch contributes mostly to over-fitting the auxiliary heads against noisy GDELT-derived targets without compensating gains on ideology classification. Under this hypothesis, the epoch-1 outperformance is not random fluctuation but the predictable interaction of dense primary supervision with noisy auxiliary supervision under fixed loss weighting; it is also one of the cleanest motivations for the adaptive-weighting future work flagged in Chapter~\ref{sec:limitations-future-work}.

Across configurations, we generally observe performance improvements after training for two epochs. The full dataset configurations show the most consistent gains, with the $T_{id}(1)$ group demonstrating substantial F1-Macro increases from 44.36--52.67 (epoch 1) to 51.40--60.78 (epoch 2), while $T_{id}(16) + T_h + T_o$ improves from 52.01 to 60.78. Similarly, the subset dataset shows improvements in several configurations, such as $T_{id}(1) + T_h + T_o$ advancing from 48.31 to 56.56. However, notable outliers exist: some configurations exhibit performance degradation or unexpectedly poor results at epoch 2, particularly evident in $T_{id}(32)$ on the subset dataset, where several variants underperform relative to epoch 1. These anomalies can be attributed to multiple factors including incomplete GDELT metadata coverage (only 35\% of the full dataset contains annotations), randomness in triplet selection during training, and potential overfitting when combining intensive triplet-loss training with limited metalinguistic supervision on the smaller dataset. The center class remains challenging across all configurations, though the best center performance emerges at epoch 2 with combined objectives, where $T_{id}(32) + T_h + T_o$ achieving 42.28 on full dataset.

\begin{table*}
\centering
\small
\begin{tabular}{@{}lcccccccc@{}}
\toprule
\textbf{Experiment} & \multicolumn{4}{c}{\textbf{Subset (1.6M, Complete GDELT)}} & \multicolumn{4}{c}{\textbf{Full (4.5M, 35\% GDELT)}} \\
\cmidrule(lr){2-5} \cmidrule(lr){6-9}
 & F1-Macro & Left & Center & Right &  F1-Macro & Left & Center & Right \\
\midrule
\multicolumn{9}{l}{\textit{Epoch 1}} \\
\midrule
$T_{id}(1)$          & 50.12 & \cellcolor{blue!15}\textbf{73.13} & 13.79 & \cellcolor{blue!15}\textbf{68.35} & 44.36 & 68.01 & 7.49 & 57.58 \\
$T_{id}(1) + T_h$    & \cellcolor{blue!15}\textbf{52.02} & 71.61 & \cellcolor{blue!15}\textbf{33.41} & 51.03 & 48.12 & 68.04 & 18.16 & 58.15 \\
$T_{id}(1) + T_o$    & 49.40 & 69.23 & 14.40 & \cellcolor{blue!15}64.57 & \textbf{52.67} & \textbf{72.26} & \textbf{24.02} & 61.74 \\
$T_{id}(1) + T_h + T_o$ & 48.31 & 69.22 & 23.06 & 52.64 & 50.21 & 68.15 & 18.49 & \textbf{63.98} \\
\midrule
$T_{id}(16)$         & 47.76 & 67.55 & 9.32 & \cellcolor{blue!15}66.41 & 51.02 & 67.99 & 16.97 & 68.09 \\
$T_{id}(16) + T_h$   & 47.89 & 67.31 & 11.85 & 64.52 & \cellcolor{blue!15}\textbf{60.55} & \textbf{74.11} & \cellcolor{blue!15}\textbf{38.22} & \cellcolor{blue!15}\textbf{69.31} \\
$T_{id}(16) + T_o$   & 53.09 & 71.48 & \cellcolor{blue!15}23.49 & 64.29 & 52.06 & 74.09 & 18.51 & 63.57 \\
$T_{id}(16) + T_h + T_o$ & \textbf{55.15} & \textbf{71.55} & \textbf{26.85} & \cellcolor{blue!15}\textbf{67.06} & 52.01 & 68.60 & 20.44 & 66.97 \\
\midrule
$T_{id}(32)$         & 52.82 & 71.76 & 25.82 & 60.88 & 56.48 & 73.23 & 24.87 & \cellcolor{blue!15}\textbf{71.35} \\
$T_{id}(32) + T_h$   & 45.73 & 65.22 & \cellcolor{blue!15}4.03 & \textbf{67.93} & \textbf{60.47} & \textbf{74.66} & \textbf{38.81} & 67.95 \\
$T_{id}(32) + T_o$   & 46.36 & 66.90 & 8.97 & 63.22 & \cellcolor{blue!15}57.18 & \cellcolor{blue!15}72.75 & \cellcolor{blue!15}34.49 & 64.30 \\
$T_{id}(32) + T_h + T_o$ & \cellcolor{blue!15}\textbf{55.59} & \cellcolor{blue!15}\textbf{72.73} & \cellcolor{blue!15}29.27 & \cellcolor{blue!15}\textbf{64.76} & 50.21 & 68.15 & 18.49 & \cellcolor{blue!15}63.98 \\
\midrule
\multicolumn{9}{l}{\textit{Epoch 2}} \\
\midrule
$T_{id}(1)$          & 50.12 & 69.28 & 18.09 & 63.00 & \textbf{56.98} & \textbf{73.82} & 33.96 & 63.55 \\
$T_{id}(1) + T_h$    & 49.59 & 72.86 & 18.19 & 57.73 & 51.40 & 68.17 & 21.81 & 64.21 \\
$T_{id}(1) + T_o$    & 53.39 & 71.65 & 28.31 & 60.20 & 55.86 & 72.46 & 31.17 & 63.93 \\
$T_{id}(1) + T_h + T_o$ & \textbf{56.56} & \textbf{72.58} & \textbf{30.39} & \textbf{66.71} & 60.78 & 73.05 & \textbf{39.48} & \textbf{69.82} \\
\midrule
$T_{id}(16)$         & 50.06 & 72.07 & 20.53 & 57.59 & 52.88 & 68.36 & 22.02 & 68.26 \\
$T_{id}(16) + T_h$   & 50.10 & 67.33 & 16.23 & \textbf{66.74} & 56.38 & \textbf{75.97} & 27.94 & 65.22 \\
$T_{id}(16) + T_o$   & 53.30 & \textbf{71.99} & 22.08 & 65.83 & 53.76 & 74.11 & 19.71 & 67.45 \\
$T_{id}(16) + T_h + T_o$ & \textbf{56.56} & 72.58 & \textbf{30.39} & 66.71 & \textbf{60.78} & 73.05 & \textbf{39.48} & \textbf{69.82} \\
\midrule
$T_{id}(32)$         & \textbf{55.41} & \textbf{73.66} & \textbf{28.34} & 64.22 & 60.15 & \textbf{73.95} & 35.88 & 70.62 \\
$T_{id}(32) + T_h$   & 46.47 & 66.05 & 03.83 & \textbf{69.52} & \textbf{62.19} & 75.97 & \textbf{39.83} & \textbf{70.76} \\
$T_{id}(32) + T_o$   & 48.74 & 67.21 & 12.83 & 66.18 & 49.73 & 69.81 & 12.68 & 66.68 \\
$T_{id}(32) + T_h + T_o$ & 48.70 & 68.01 & 15.57 & 62.52 & 59.33 & 72.66 & \textbf{42.28} & 63.06 \\
\bottomrule
\end{tabular}
\caption{Continued pre-training results on both GDELT limited subset and full \textsc{BigNewsBLN} dataset, evaluated after epoch 1 (top) and epoch 2 (bottom). \colorbox{blue!15}{Highlighted} cells indicate epoch~1 outperforms epoch~2 for that configuration and split. Bold indicates best result per column within each epoch block.}
\label{tab:pretraining-combined-full}
\end{table*}
