\chapter{Community-Level ATE Results}
\label{sec:appendix-community-ate}

This appendix provides the full community-level ATE breakdown for the nine communities identified in Section~\ref{sec:community-level-results}. Four communities (0, 1, 2, and 5) produce at least one significant ATE across the five annotation paradigms, with Community~2 (Public Health) achieving full cross-paradigm replication.

\section{Complete Community-Level ATE Table}

Table~\ref{tab:community_ate_full} presents the average treatment effect of community-level sentiment on ideology classification across all nine communities that met the minimum presence threshold. We compare estimates derived from human annotations against three LLM-based annotation pipelines (GPT-4o-mini (baseline), Fine-tuned GPT-4o-mini, and Llama-3.3-70B) and one fine-tuned encoder model (Fine-tuned RoBERTa). Community theme labels correspond to the thematic groupings in Table~\ref{tab:community-structure}.

\begin{table}[h!]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{l c c c c c c}
  \toprule
  Community & Human & Fine-tuned RoBERTa & GPT-4o-mini (baseline) & Fine-tuned GPT-4o-mini & Llama-3.3-70B & Num.\ Articles\\
  \midrule
  Comm.\ 0 (White House/Congress) & $\mathbf{-0.024^{*}}\ [-0.040,\ -0.005]$ & $\mathbf{-0.023^{*}}\ [-0.039,\ -0.004]$ & $-0.013\ [-0.028,\ +0.002]$ & $\mathbf{-0.025^{*}}\ [-0.043,\ -0.007]$ & $-0.008\ [-0.027,\ +0.009]$ & 2473\\
  Comm.\ 1 (Elections/Campaigns) & $+0.016\ [-0.002,\ +0.034]$ & $+0.011\ [-0.008,\ +0.028]$ & $\mathbf{+0.034^{*}}\ [+0.019,\ +0.050]$ & $+0.009\ [-0.010,\ +0.026]$ & $\mathbf{+0.056^{*}}\ [+0.037,\ +0.073]$ & 3628\\
  Comm.\ 2 (Public Health) & $\mathbf{-0.036^{*}}\ [-0.053,\ -0.019]$ & $\mathbf{-0.027^{*}}\ [-0.045,\ -0.010]$ & $\mathbf{-0.035^{*}}\ [-0.050,\ -0.021]$ & $\mathbf{-0.031^{*}}\ [-0.048,\ -0.013]$ & $\mathbf{-0.045^{*}}\ [-0.063,\ -0.029]$ & 1116\\
  Comm.\ 3 (Media/Technology) & $+0.003\ [-0.017,\ +0.020]$ & $-0.003\ [-0.023,\ +0.016]$ & $-0.003\ [-0.019,\ +0.014]$ & $-0.000\ [-0.017,\ +0.018]$ & $+0.006\ [-0.014,\ +0.025]$ & 890\\
  Comm.\ 4 (Culture/Education) & $+0.013\ [-0.014,\ +0.025]$ & $+0.007\ [-0.015,\ +0.019]$ & $+0.007\ [-0.011,\ +0.021]$ & $+0.006\ [-0.018,\ +0.021]$ & $+0.012\ [-0.013,\ +0.025]$ & 720\\
  Comm.\ 5 (Env./Fiscal Policy) & $-0.001\ [-0.017,\ +0.017]$ & $-0.003\ [-0.019,\ +0.015]$ & $\mathbf{-0.019^{*}}\ [-0.030,\ -0.003]$ & $-0.005\ [-0.021,\ +0.011]$ & $-0.010\ [-0.026,\ +0.009]$ & 1480\\
  Comm.\ 6 (Constitutional Rights) & $-0.007\ [-0.026,\ +0.008]$ & $-0.000\ [-0.020,\ +0.014]$ & $+0.011\ [-0.009,\ +0.026]$ & $-0.000\ [-0.020,\ +0.014]$ & $+0.011\ [-0.011,\ +0.028]$ & 1094\\
  Comm.\ 8 (Foreign Policy) & $+0.013\ [-0.004,\ +0.029]$ & $+0.009\ [-0.011,\ +0.023]$ & $+0.000\ [-0.011,\ +0.013]$ & $+0.009\ [-0.008,\ +0.024]$ & $+0.011\ [-0.006,\ +0.026]$ & 2679\\
  Comm.\ 9 (Criminal Justice) & $+0.001\ [-0.017,\ +0.017]$ & $+0.005\ [-0.014,\ +0.021]$ & $+0.009\ [-0.007,\ +0.024]$ & $+0.009\ [-0.011,\ +0.024]$ & $+0.005\ [-0.014,\ +0.021]$ & 1364\\
  \bottomrule
  \end{tabular}%
  }
  \caption{Community-level ATE by annotation paradigm (continuous sentiment, min presence${}=50$, min weight${}=2$). $^*p < 0.05$.}
  \label{tab:community_ate_full}
\end{table}

Four of nine communities produce at least one significant ATE, yielding 14 significant cells in total. This represents a substantial increase in statistical power relative to prior runs with smaller datasets, where only two cells reached significance. Community~2 (Public Health) is the standout result: all five paradigms produce significant negative ATEs (Human $-0.036$, Fine-tuned RoBERTa $-0.027$, baseline GPT $-0.035$, Fine-tuned GPT $-0.031$, Llama $-0.045$), making it the only community with full cross-paradigm replication, analogous to the partisan topics (e.g.\ Healthcare, Democratic Party) that achieve universal significance at the topic level in Table~\ref{tab:multi-continuous-full}.

Community~0 (White House and Congress) produces significant negative ATEs under three paradigms (Human $-0.024$, Fine-tuned RoBERTa $-0.023$, Fine-tuned GPT $-0.025$); baseline GPT ($-0.013$) and Llama ($-0.008$) agree directionally but do not reach significance. Community~1 (Elections and Campaigns) exhibits the opposite pattern: only the two zero-shot paradigms reach significance (baseline GPT $+0.034$, Llama $+0.056$), both in the positive (rightward) direction, while Human ($+0.016$), Fine-tuned RoBERTa ($+0.011$), and Fine-tuned GPT ($+0.009$) remain non-significant. Community~5 (Environment and Fiscal Policy) produces a single significant cell under baseline GPT ($-0.019$); all other paradigms estimate small negative effects that span zero.

Directional agreement across paradigms is high. In Communities~0, 2, and~5 all five paradigms agree on a negative direction, and in Community~1 all five agree on a positive direction. The remaining communities (3, 4, 6, 8, 9) show small, non-significant effects with occasional directional splits, consistent with the absence of a strong partisan signal in their constituent topics.
