\chapter{Human Annotation Tests}
\label{sec:appendix-human-tests}

To establish a reference point for system performance, we conducted an informal human evaluation study using a randomly sampled subset of 102 articles from the AllSides dataset Media test split. The articles were stratified equally across the three political ideology categories (left, center, right), with 34 articles representing each class.

\begin{table}[ht]
\centering
\small
\begin{tabular}{lrrrr}
\toprule
\textbf{Annotator}  & \textbf{F1-Macro} & \textbf{Left} & \textbf{Center} & \textbf{Right} \\
\midrule
Annotator 1 & 46.57 & 43.59 & 41.27 & 54.84 \\
Annotator 2 & 35.46 & 39.02 & 26.67 & 40.68 \\
Annotator 3 & 40.96 & 23.26 & 56.14 & 43.48 \\
Annotator 4 & 44.75 & 43.37 & 32.26 & 58.62 \\
\midrule
Average & 41.94 & 37.31 & 39.09 & 49.41 \\
\midrule
Kripp. $\alpha$ & 0.241 & 0.229 & 0.188 & 0.345 \\
\bottomrule
\end{tabular}
\caption{Annotator Performance}
\label{tab:human_baseline}
\end{table}

Four annotators participated in the evaluation process. All annotators were under 35 years of age, self-identified as politically liberal to moderate, and reported good familiarity with U.S. politics. The participants included three men and one woman, all of whom had resided in the U.S for more than 10 years. They were all volunteers who were acquainted socially with the authors, received no compensation, and were informed that anonymised results may be used for publication. 

Each annotator reviewed the complete set of articles and assigned political ideology classifications (left, lean-left, center, lean-right, right) based solely on article content, without explicit access to publication source information. Although annotators were allowed to rate across five categories, the results were aggregated into three ideologies (left and lean-left were grouped into left, right and lean-right were grouped into right) for ease of analysis.

The experiment was conducted over a one-month period in a non-controlled environment, allowing annotators to complete the task at their own pace. Annotators were given a Google Sheet (sample available at \url{https://tinyurl.com/yv83wfrx}) containing article content, along with columns to record their classifications. The instructions were as follows:

\begin{quote}
In the articles tab, you will find 102 articles. Please read as much as you need to classify them as having one of the following biases:
\begin{itemize}
    \item \href{https://www.allsides.com/media-bias/left}{left}
    \item \href{https://www.allsides.com/media-bias/left-center}{lean-left}
    \item \href{https://www.allsides.com/media-bias/center}{center}
    \item \href{https://www.allsides.com/media-bias/right-center}{lean-right}
    \item \href{https://www.allsides.com/media-bias/right}{right}
\end{itemize}

The definitions of these biases are linked above, you can use these to inform your intuition. If you are curious: forms of media bias can be found \href{https://www.allsides.com/media-bias/how-to-spot-types-of-media-bias}{here}. This will likely take between 1-2 hours.

In the demographics tab, there is a brief demographics questionnaire. Please do fill it out.

Please send me your responses by July 2! Earlier is better :)
\end{quote}

\paragraph{Annotator-pool biases and study-design caveats.} The annotator pool described above introduces several plausible sources of bias that should be disclosed transparently rather than buried in the methodology, because they directly bear on how the human F1 = 41.94 number can and cannot be interpreted. First, all four annotators were socially acquainted with the author, which creates a social-desirability deference risk: annotators may have shaped their behavior around perceived expectations of the study even with no explicit cue to do so. Second, the pool was uncompensated and worked at their own pace over a one-month window in a non-controlled environment, which permits fatigue, interruption, and revisiting of articles mid-protocol in ways that controlled annotation studies typically prevent. Third, all four annotators self-identified as politically liberal to moderate; an ideologically homogeneous pool can produce out-group salience effects, where right-leaning content is paradoxically easier to recognize as right-leaning precisely because it sits outside the annotators' own political vantage point (this is consistent with the per-class F1 numbers in Table~\ref{tab:human_baseline}, where all four annotators score highest on the Right class). Fourth, the protocol was self-administered, with no enforced article ordering or attention checks. Each of these issues would be addressed in a controlled annotation study; this study is not one. The headline number F1 = 41.94 should therefore be read as an informal lay-annotator reference point under these specific conditions, not as a directly comparable replacement for trained-expert annotation studies such as MBIC or BABE.

Despite relatively strong demographic homogeneity among annotators, we observed substantial variability in judgment as evidenced by relatively low inter-annotator agreement (Krippendorff's $\alpha = 0.241$) and consistently moderate F-1 metrics across all participants (Table~\ref{tab:human_baseline}). These findings suggest that political ideology classification presents significant challenges even for human evaluators. This difficulty likely stems from the inherently subjective nature of ideology perception, as well as from subtle variations in annotators' interpretative frameworks and ideological calibrations. Anecdotally, annotators reported that this task was particularly challenging for non-hyperpartisan samples, and indicated that source information would have increased their confidence, suggesting a general human instinct to intertwine text and source bias. This anecdotal report converges with a quantitative finding from the model side: the same Center class on which annotators report the most difficulty is also the class on which our best models struggle hardest (39.83 F1 on Center for the metalinguistic RoBERTa). The convergence is informative: it suggests that center-class difficulty is task-inherent (non-hyperpartisan content carries less marked ideological signal), rather than a model-specific artifact, and motivates treating Center performance as a structural rather than incidental challenge.
