\chapter{Community-Level Mediation Results}
\label{sec:appendix-community-mediation}

This appendix provides the full community-level mediation results referenced in Section~\ref{sec:community-mediation}, including pathway selection rationale and complete numerical tables.

\section{Pathway Selection}

Six treatment-mediator pairs were evaluated across communities $\{3, 9, 11\}$, but only $9 \to 11$ and $11 \to 9$ yielded non-degenerate mediation decompositions. The per-pathway outcome is:

\begin{itemize}
    \item $3 \to 9$: 0 articles co-occur, pathway eliminated.
    \item $3 \to 11$: 0 articles co-occur, pathway eliminated.
    \item $9 \to 3$: $< 50$ co-occurring articles, drops below the minimum-presence threshold.
    \item $11 \to 3$: $< 50$ co-occurring articles, drops below the minimum-presence threshold.
    \item $9 \to 11$: 139 articles co-occur, pathway retained.
    \item $11 \to 9$: 772 articles co-occur, pathway retained.
\end{itemize}

The six-pair matrix therefore reduces to two unique, well-specified analyses, both gated by the topic co-occurrence filter rather than by any modelling failure.

\section{Full Mediation Results}

Table~\ref{tab:mediation_full} reports total effects (TE), natural direct effects (NDE), and natural indirect effects (NIE) for both viable pathways under the $Q_{25} \to Q_{75}$ contrast across all available annotation paradigms.

\begin{table}[h]
\centering
\small
\begin{tabular}{lcccccc}
\toprule
Annotator & TE & 95\% CI & NDE & 95\% CI & NIE & \% med. \\
\midrule
\multicolumn{7}{l}{\textit{Community $9 \to 11$ ($N=139$, sentiment contrast $-0.3$ to $0.0$)}} \\
\midrule
Human              & $-0.045$ & $[-0.164,\ +0.075]$ & $-0.059$ & $[-0.151,\ +0.045]$ & $+0.015$ & --- \\
Fine-tuned RoBERTa & $-0.069$ & $[-0.181,\ +0.043]$ & $-0.080$ & $[-0.147,\ +0.031]$ & $+0.011$ & --- \\
Fine-tuned GPT     & $\mathbf{-0.108^*}$ & $\mathbf{[-0.208,\ -0.008]}$ & $\mathbf{-0.105^*}$ & $\mathbf{[-0.170,\ 0.000]}$ & $-0.003$ & 3.2\% \\
GPT (baseline)     & $-0.059$ & $[-0.141,\ +0.023]$ & $-0.032$ & $[-0.102,\ +0.033]$ & $-0.026$ & --- \\
Llama-3.3-70B      & $-0.007$ & $[-0.148,\ +0.134]$ & $+0.023$ & $[-0.080,\ +0.099]$ & $-0.030$ & --- \\
\midrule
\multicolumn{7}{l}{\textit{Community $11 \to 9$ ($N=772$, sentiment contrast $-0.4$ to $0.1$)}} \\
\midrule
Human              & $-0.045$ & $[-0.162,\ +0.071]$ & $-0.049$ & $[-0.100,\ +0.012]$ & $+0.004$ & --- \\
Fine-tuned RoBERTa & $-0.016$ & $[-0.133,\ +0.102]$ & $-0.015$ & $[-0.079,\ +0.034]$ & $-0.000$ & --- \\
Fine-tuned GPT     & $+0.056$ & $[-0.069,\ +0.181]$ & $+0.069$ & $[0.000,\ +0.116]$  & $-0.013$ & --- \\
GPT (baseline)     & $+0.007$ & $[-0.097,\ +0.111]$ & $+0.009$ & $[-0.042,\ +0.059]$ & $-0.002$ & --- \\
Llama-3.3-70B      & $+0.008$ & $[-0.114,\ +0.130]$ & $+0.012$ & $[-0.046,\ +0.075]$ & $-0.004$ & --- \\
\bottomrule
\end{tabular}
\caption{Community-level mediation results ($Q_{25} \to Q_{75}$ contrast). $^*$Significant at 95\% (CI excludes zero).}
\label{tab:mediation_full}
\end{table}

For the $9 \to 11$ pathway, Fine-tuned GPT-4o-mini is the only annotator to produce a significant total effect (TE $= -0.108$), with nearly all of it attributable to the direct pathway (3.2\% mediated). Human and baseline GPT annotators show point estimates in the same negative direction but with wide, zero-crossing intervals. Fine-tuned RoBERTa (TE $= -0.069$) ranks second in magnitude after Fine-tuned GPT-4o-mini, further supporting the directional consensus. Llama-3.3-70B produces a near-zero TE ($-0.007$) with the widest confidence interval, consistent with its general attenuation pattern observed in the ATE analysis. All five annotators are directionally consistent (negative), suggesting the non-significance of human, RoBERTa, baseline GPT, and Llama reflects power limitations at $N=139$ rather than directional disagreement.

For the $11 \to 9$ pathway, no annotator reaches significance. The Fine-tuned GPT-4o-mini effect reverses sign relative to the $9 \to 11$ direction ($+0.056$ vs.\ $-0.108$), reinforcing the asymmetry interpretation: Community~9's effect on ideology is concentrated and direct, while Community~11's influence is diffuse across many mediators. Llama-3.3-70B produces a small positive TE ($+0.008$), again attenuated relative to other paradigms. Human and Fine-tuned RoBERTa are the only annotators to produce negative TEs for this pathway ($-0.045$ and $-0.016$, respectively), while all three LLM-based paradigms estimate positive effects, reinforcing the pattern of RoBERTa tracking human directional tendencies.

\section{Full Multi-Treatment DML Results (Continuous Encoding)}

Table~\ref{tab:multi-continuous-full} reports the complete multi-treatment DML results summarized in Table~\ref{tab:multi-continuous} of Chapter~\ref{ch:causal-inference}, including the 95\% bootstrap confidence interval ($B = \num{2000}$) accompanying every ATE estimate.

\begin{sidewaystable}
  \centering
  \sisetup{table-format=+1.4}
  \small
  \begin{tabular}{l c
      S[table-format=+1.4] @{~} l @{~} l
      S[table-format=+1.4] @{~} l @{~} l
      S[table-format=+1.4] @{~} l @{~} l
      S[table-format=+1.4] @{~} l @{~} l
      S[table-format=+1.4] @{~} l @{~} l}
    \toprule
    & & \multicolumn{3}{c}{\textbf{Human}}
    & \multicolumn{3}{c}{\textbf{Fine-tuned RoBERTa}}
    & \multicolumn{3}{c}{\textbf{GPT-4o-mini (baseline)}}
    & \multicolumn{3}{c}{\textbf{Fine-tuned GPT-4o-mini}}
    & \multicolumn{3}{c}{\textbf{Llama-3.3-70B}} \\
    \cmidrule(lr){3-5}\cmidrule(lr){6-8}\cmidrule(lr){9-11}\cmidrule(lr){12-14}\cmidrule(lr){15-17}
    \textbf{Topic} & {\textbf{N}}
      & {\textbf{ATE}} & \multicolumn{2}{c}{\textbf{95\% CI}}
      & {\textbf{ATE}} & \multicolumn{2}{c}{\textbf{95\% CI}}
      & {\textbf{ATE}} & \multicolumn{2}{c}{\textbf{95\% CI}}
      & {\textbf{ATE}} & \multicolumn{2}{c}{\textbf{95\% CI}}
      & {\textbf{ATE}} & \multicolumn{2}{c}{\textbf{95\% CI}} \\
    \midrule
    Hillary Clinton & 51
      & +0.0051 & {[}${-0.019}$ & ${+0.018}${]}
      & \textbf{+0.0220} & {[}${+0.001}$ & ${+0.036}${]}
      & -0.0064 & {[}${-0.020}$ & ${+0.010}${]}
      & -0.0004 & {[}${-0.015}$ & ${+0.023}${]}
      & -0.0034 & {[}${-0.013}$ & ${+0.020}${]} \\
    Media Bias & 115
      & +0.0024 & {[}${-0.024}$ & ${+0.015}${]}
      & +0.0203 & {[}${-0.001}$ & ${+0.034}${]}
      & -0.0130 & {[}${-0.021}$ & ${+0.010}${]}
      & -0.0191 & {[}${-0.025}$ & ${+0.010}${]}
      & +0.0057 & {[}${-0.011}$ & ${+0.026}${]} \\
    FBI & 72
      & +0.0182 & {[}${-0.010}$ & ${+0.029}${]}
      & -0.0052 & {[}${-0.022}$ & ${+0.013}${]}
      & -0.0006 & {[}${-0.019}$ & ${+0.009}${]}
      & +0.0115 & {[}${-0.011}$ & ${+0.032}${]}
      & -0.0077 & {[}${-0.023}$ & ${+0.012}${]} \\
    Donald Trump & 233
      & -0.0089 & {[}${-0.028}$ & ${+0.010}${]}
      & -0.0025 & {[}${-0.024}$ & ${+0.012}${]}
      & -0.0112 & {[}${-0.025}$ & ${+0.004}${]}
      & +0.0011 & {[}${-0.015}$ & ${+0.022}${]}
      & -0.0087 & {[}${-0.025}$ & ${+0.009}${]} \\
    Media Watch & 58
      & -0.0032 & {[}${-0.027}$ & ${+0.017}${]}
      & -0.0079 & {[}${-0.030}$ & ${+0.015}${]}
      & +0.0093 & {[}${-0.006}$ & ${+0.023}${]}
      & +0.0078 & {[}${-0.007}$ & ${+0.026}${]}
      & -0.0104 & {[}${-0.024}$ & ${+0.006}${]} \\
    Russia Probe & 57
      & +0.0011 & {[}${-0.021}$ & ${+0.026}${]}
      & +0.0045 & {[}${-0.023}$ & ${+0.026}${]}
      & -0.0002 & {[}${-0.017}$ & ${+0.016}${]}
      & -0.0210 & {[}${-0.043}$ & ${+0.008}${]}
      & +0.0100 & {[}${-0.018}$ & ${+0.030}${]} \\
    Race and Racism & 104
      & -0.0013 & {[}${-0.020}$ & ${+0.017}${]}
      & +0.0040 & {[}${-0.018}$ & ${+0.019}${]}
      & -0.0041 & {[}${-0.010}$ & ${+0.011}${]}
      & +0.0050 & {[}${-0.009}$ & ${+0.022}${]}
      & +0.0067 & {[}${-0.004}$ & ${+0.021}${]} \\
    Defense and Security & 76
      & +0.0017 & {[}${-0.027}$ & ${+0.022}${]}
      & +0.0034 & {[}${-0.015}$ & ${+0.022}${]}
      & -0.0076 & {[}${-0.032}$ & ${+0.006}${]}
      & +0.0112 & {[}${-0.022}$ & ${+0.025}${]}
      & +0.0068 & {[}${-0.022}$ & ${+0.021}${]} \\
    LGBTQ Issues & 73
      & +0.0027 & {[}${-0.013}$ & ${+0.019}${]}
      & +0.0021 & {[}${-0.014}$ & ${+0.023}${]}
      & -0.0045 & {[}${-0.013}$ & ${+0.010}${]}
      & -0.0144 & {[}${-0.024}$ & ${+0.006}${]}
      & -0.0063 & {[}${-0.022}$ & ${+0.008}${]} \\
    Politics & 594
      & -0.0035 & {[}${-0.021}$ & ${+0.014}${]}
      & -0.0023 & {[}${-0.018}$ & ${+0.018}${]}
      & +0.0098 & {[}${-0.006}$ & ${+0.022}${]}
      & \textbf{+0.0217} & {[}${+0.003}$ & ${+0.034}${]}
      & +0.0017 & {[}${-0.013}$ & ${+0.019}${]} \\
    Economy and Jobs & 56
      & -0.0001 & {[}${-0.018}$ & ${+0.020}${]}
      & +0.0019 & {[}${-0.018}$ & ${+0.020}${]}
      & -0.0096 & {[}${-0.023}$ & ${+0.002}${]}
      & -0.0095 & {[}${-0.025}$ & ${+0.009}${]}
      & +0.0083 & {[}${-0.013}$ & ${+0.024}${]} \\
    \bottomrule
  \end{tabular}
  \caption{Full multi-treatment DML results: ATE per topic under continuous encoding (one-SD increase in sentiment score) with 95\% bootstrap CIs ($B = \num{2000}$). Bolded ATEs have CIs that exclude zero.}
  \label{tab:multi-continuous-full}
\end{sidewaystable}

\section{Confounder Reduction Details}

Confounders consist of all community sentiment scores except those of the treatment and mediator communities, reduced in two sequential steps:

\begin{enumerate}
    \item \textbf{Presence filter:} Communities whose sentiment is non-zero in fewer than 50 articles in the treatment subsample are dropped.
    \item \textbf{PCA:} The remaining matrix is reduced via PCA retaining 95\% of variance.
\end{enumerate}

Because this reduction depends only on the treatment subsample, not on ideology labels, it is identical across all annotation paradigms for a given treatment community. All treatment communities begin with 11 potential confounders. For Communities 1, 3, and 11, the presence filter removes 3, leaving 8; PCA further reduces these to 7 components. Community~9 loses one additional community (4 removed, leaving 7), and PCA retains all 7. All configurations therefore yield 7 final confounder components; the choice of mediator community has no effect on this reduction.
