
\section{Baseline Performance Analysis}
\label{sec:baseline-results}

Table~\ref{tab:main_baseline} presents comprehensive baseline results across all evaluated systems. The performance patterns reveal several important findings regarding the challenges and opportunities in automated political bias classification.

\paragraph{Threat to validity: LLM training-data contamination.} Before interpreting any of the LLM numbers below, it is important to flag the central threat to their validity: the LLMs evaluated here were trained on web-scale corpora that very plausibly include AllSides articles, AllSides-derived datasets, or close paraphrases thereof. This means an apparently competitive zero-shot F1 may reflect memorized source--ideology pairings rather than learned content-based ideology, and headline F1 alone cannot distinguish the two. This is not a footnote concern: it is the reason we treat zero-shot LLM accuracy on this task with caution throughout, and it is the gap that the causal-probing analysis in Chapter~\ref{sec:causal-inference} is designed to diagnose by asking whether LLM and human annotators rely on the same causal pathways from sentiment and topic to ideology.

\subsection{Source Leakage Effects}

The comparison between Media Split and Random Split performance provides crucial insights into source leakage susceptibility. All transformer models demonstrate substantially higher performance on the Random Split, with improvements of 30--50 F1 points across most configurations. This performance differential is most pronounced for center-category classification, where Random Split performance exceeds Media Split by over 40 points in several cases.

A plausible explanation is that center articles are simply harder content: the underlying ideological signal is more diffuse and less lexically marked than at either pole, so when source identity \emph{is} available the model leans on it especially heavily for the center class because that is the path of least resistance. The substantial performance degradation on the Media Split confirms that those source-derived shortcuts do not correspond to genuine ideological content markers, highlighting the importance of source-controlled evaluation for political bias detection systems.

\subsection{Transformer Architecture and Large Language Model Performance}

Among general-purpose transformers, RoBERTa achieved the highest Media Split performance (F1-Macro = 42.34), followed by BERT (38.26) and BART (36.01). However, all general-purpose models were substantially outperformed by the domain-adapted POLITICS system (57.66), demonstrating the value of triplet-loss pre-training in learning useful embeddings for this kind of nuanced classification task.

Zero-shot LLM performance varied considerably across model families and sizes. GPT family models achieved competitive results, with o3-mini reaching 56.92 F1-Macro—approaching the performance of the specialized POLITICS system. Notably, LLMs demonstrated relatively stronger performance on center (53.46) and right (70.78) categories compared to transformer baselines. The o3-mini lead over the other GPT variants is consistent with its design: o3-mini is an explicitly reasoning-oriented model, and a directional ideology call (which requires weighing topical emphasis, framing, and stance against each other rather than simply pattern-matching keywords) is exactly the kind of decomposable judgment task on which reasoning-tuned models would be expected to outperform their non-reasoning peers.

LLaMA family models, in contrast, underperformed significantly, with both evaluated models falling below human baseline performance. We tested multiple prompt variants on the LLaMA models without observing meaningful improvement, which lets us rule out prompt sensitivity as the dominant cause. The remaining candidates are training-data composition (less English-language political news in the LLaMA training mix, or different curation choices around polarized content) and instruction-following capacity at this directional-classification task; we commit to those two as the most plausible mechanisms for the LLaMA underperformance gap.

\subsection{Performance Interpretation}

The contamination concern flagged at the top of this section is the dominant interpretation challenge for the LLM rows of Table~\ref{tab:main_baseline}: it does not apply to our transformer baselines, where training data composition is fully controlled and transparent, so transformer Media-split results remain the most reliable assessment of content-based classification performance. The causal-probing analysis in Chapter~\ref{sec:causal-inference} is the diagnostic that resolves the LLM contamination question.

These baseline findings align with observations by \citet{Wen_2023}, who demonstrate that while LLMs achieve competitive performance on explicit classification tasks, they consistently underperform specialized fine-tuned models on nuanced tasks requiring deeper contextual understanding.

\subsection{Undersampling Analysis}

All fine-tuning experiments in this thesis, both the transformer baselines reported here and the continued pre-training experiments in Section~\ref{sec:continued-pretraining}, are conducted without class-balancing undersampling. A preliminary study (see Appendix~\ref{appendix:undersampling}) shows that undersampling does not yield consistent improvements: effects are highly model-dependent, with most models degrading overall when undersampling is applied. Where undersampling does help, gains tend to accrue to classes that are already relatively well-classified (particularly Left), rather than to the harder Center and Right categories that most challenge these models. Because Chapter~\ref{sec:continued-pretraining}'s continued pre-training is specifically aimed at improving the harder Center and Right categories, we avoid undersampling here, which would trade those gains away. We therefore opt for standard fine-tuning without undersampling to avoid inadvertently trading off performance on the most difficult classes.
