

\section{Large Language Model Evaluation}
\label{sec:llm-baselines}

The recent emergence of large-scale language models necessitates comprehensive evaluation of their capabilities for political bias classification. We assess five recent large language models from the GPT \citep{openai2024gpt4,openai_o3_systemcard_2025,openai_gpt5_systemcard_2025} and LLaMA \citep{touvron2023llama} families, spanning different scales, organizations, and release dates, in zero-shot configuration to establish baseline performance without task-specific adaptation.

\subsection{Model Selection and Access}

Our evaluation encompasses models from two prominent LLM families. From the GPT family, we evaluate GPT-4o-mini, GPT-5-nano, and o3-mini, all accessed via the OpenAI API. From the LLaMA family, we evaluate LLaMA-3.1-8B-instant and LLaMA-3.3-70B-versatile, accessed via the GroqCloud API platform. Both LLaMA and GPT models were evaluated at temperature 1.0. For computational efficiency, we conducted preliminary screening on a 100-article subset, ranked candidate GPT models by F1-Macro on that subset, and selected the top three for full-scale evaluation.

\subsection{Prompt Design and Evaluation Protocol}

To ensure consistent evaluation across models, we employed a standardized prompt format designed to elicit structured responses:

\begin{quote}
\texttt{You are a political bias classifier specific to U.S. politics. Given the user's INPUT TEXT, return only valid JSON of the form \{''bias'':''left|center|right''\}. No extra text.}
\end{quote}

This prompt design minimizes variability in response format while providing clear task specification. All models were evaluated exclusively on the Media Split configuration to maintain consistency with our content-based evaluation framework. We also tested multiple prompt variants, including chain-of-thought and detailed definition prompts, but found no significant performance improvements over the baseline prompt; the full per-variant comparison is reported in the appendix table referenced in Appendix~\ref{appendix:opro}.

