\section{Fine-tuned Transformer Architectures}
\label{sec:transformer-baselines}

Our transformer baseline evaluation encompasses both general-purpose pre-trained language models and domain-adapted architectures specifically designed for political text analysis. This comparison enables assessment of the value of domain-specific adaptation for political bias detection.

\subsection{General-Purpose Transformer Models}

We evaluated three representative pre-trained transformer architectures: BERT \citep{devlin2019bert}, RoBERTa \citep{liu2019roberta}, and BART \citep{lewis2020bart}. Each model was fine-tuned end-to-end for three-class political bias classification by adding a linear classification head to the pre-trained encoder representations.

All models employed identical training procedures to ensure fair comparison. Fine-tuning utilized the AdamW optimizer with learning rate scheduling and early stopping based on validation set performance. Complete hyperparameter configurations and training details are provided in Table~\ref{tab:finetuning_config} in the appendices.

\subsection{Domain-Adapted Architecture: POLITICS Model}

We include the current state-of-the-art system for article-level political bias classification, the POLITICS model developed by \citet{liu2022politics}. POLITICS is a RoBERTa-based encoder enhanced through continued pre-training using a contrastive triplet-loss framework. The intuition behind the design is straightforward in plain language: by repeatedly pulling together ideologically similar articles drawn from \emph{different} sources, the contrastive objective denies the encoder the ability to use outlet identity as a shortcut, forcing it to recover ideology from content. This is precisely the mechanism by which contrastive pre-training defeats source leakage. Triplets are generated with the following sampling procedure:

\begin{itemize}
    \item \textbf{Anchor}: Article $a$ with political label $l \in \{\text{left}, \text{right}\}$
    \item \textbf{Positive}: Article $p$ with identical label $l$ from any source
    \item \textbf{Negative}: Article $n$ with different label $m \neq l$ from any source
\end{itemize}

Note that anchors are sampled only from $\{\text{left}, \text{right}\}$, never from the center class. Center articles are not used as anchors but are still learned: by being repeatedly pushed away from both left and right anchors during contrastive training, the center class is shaped as the region of embedding space that is jointly distant from both partisan poles, rather than as a class with its own positive supervision signal.

The triplet loss objective encourages the model to produce representations where ideologically similar articles from different sources are closer in embedding space than ideologically different articles:

\begin{equation}
    \mathcal{L}_{\text{triplet}} = \sum_{\langle a,p,n \rangle \in \mathcal{T}} \max\left( \|f(a) - f(p)\|_2 - \|f(a) - f(n)\|_2 + \delta, 0\right)
    \label{eqn:triplet_loss}
\end{equation}

where $f(\cdot)$ denotes the model's [CLS] token representation, and $\delta$ is a margin hyperparameter. The POLITICS model incorporates both \emph{story-level} and \emph{ideology-level} objectives during continued pre-training, and the two play distinct roles. The story-level objective treats articles covering the same underlying event as positive pairs, which encourages the encoder to abstract over topical surface (the same event reported by different outlets should land near each other regardless of word choice) and prevents topic from being memorized as a stand-in for ideology. The ideology-level objective treats articles sharing a political orientation as positive pairs, which encourages clustering on ideological orientation specifically. Together they prevent the most pernicious shortcut available to a naive classifier, using \emph{topic} as a proxy for ideology, by forcing topic-invariance and ideology-clustering to coexist in the same representation space.
