\section{Methodology and Architecture}
\label{sec:methodology}

The research question this chapter addresses is: do auxiliary metalinguistic objectives improve over a pure contrastive pre-training baseline for article-level political ideology classification, and if so, under what conditions (which auxiliary signals, what sampling rate, what corpus coverage)? We formulate the integration of metalinguistic features as a multi-task learning problem where the primary political bias classification objective is augmented with auxiliary tasks that capture complementary aspects of political discourse. The primary task is political ideology classification using contrastive triplet-loss objectives. This is supplemented by two auxiliary tasks: multi-label topic classification capturing thematic focus, and regression of sentiment polarity reflecting tonal characteristics. This formulation enables the model to learn representations that are sensitive to both explicit ideological content and implicit stylistic markers that correlate with political orientation.

Our technical approach builds upon the RoBERTa transformer architecture enhanced through continued pre-training with multiple supervised objectives. 
\subsection{Metalinguistic Feature Sources}
\label{sec:feature-sources}

To obtain large-scale metalinguistic annotations without prohibitive manual labeling costs, we leverage the Global Database of Events, Language, and Tone (GDELT) Project,\footnote{\url{https://www.gdelt.org/}} a comprehensive news monitoring system that processes articles across 100+ languages and extracts structured metadata through automated NLP pipelines. We note that GDELT's V2Themes and V2Tone are produced by automated NLP pipelines whose precise implementation details are not publicly documented, and which therefore carry unknown but non-trivial error rates. Our auxiliary supervision is consequently noisy supervision: any auxiliary head is being asked to predict a label that is itself the output of a separate, imperfect classifier. This places a structural ceiling on how much an auxiliary objective can contribute (the heads can never be more accurate than the GDELT pipelines they are supervised against), and is part of how we interpret the magnitude of the metalinguistic gains reported below. Following \citet{ronnback2025biasoutlets}, who demonstrate the effectiveness of GDELT-derived features for source-level ideology detection, we integrate two complementary metadata streams:

\begin{description}
    \item[\textsc{V2Themes}:] Multi-label topic annotations generated through lexicon-based keyword matching against a curated taxonomy of 2,300+ categories (e.g., \texttt{ENV\_CLIMATECHANGE}, \texttt{ECON\_INFLATION}). Each article receives multiple theme codes reflecting its topical composition, enabling models to learn associations between ideological orientation and topic selection patterns.
    
    \item[\textsc{V2Tone}:] Sentiment polarity scores computed as the mean valence of matched lexical items, yielding continuous values typically in the range $[-10, +10]$. While GDELT provides six tonal subdimensions (overall tone, positive/negative scores, polarity, activity reference density, and self-reference density), we employ only the aggregate tone score to maintain computational efficiency.
\end{description}

These features capture complementary aspects of political discourse: thematic annotations reflect editorial choices about which topics receive coverage and emphasis, while tonal measurements quantify the sentiment orientation of the presentation.

Our pre-training dataset consists of articles from the \textsc{BigNewsBLN} dataset \citep{liu2022politics} for which GDELT metadata is available. This subset comprises approximately 1.6 million articles containing both \textsc{V2Themes} and \textsc{V2Tone} annotations, or around 35\% of the total \textsc{BigNewsBLN} dataset. This extended dataset is available 
\href{https://huggingface.co/datasets/dragonslayer631/bignewsalign-with-gdelt}{on HuggingFace} 
.\footnote{The data is available under the CC-BY-SA 4.0 license, same as the original dataset \citep{liu2022politics}.}

\subsection{Multi-Task Learning Framework}
\label{sec:multitask-framework}

We implement a unified multi-task learning architecture that optimizes four concurrent objectives during continued pre-training:

\begin{align}
\mathcal{L} &=  \mathcal{L}_{trip} + \mathcal{L}_{T_h} + \mathcal{L}_{T_o} + \mathcal{L}_{MLM} \label{eq:total-loss}
\end{align}

The four component losses are:

\begin{description}
    \item[\textbf{Ideological Representation Loss} ($\mathcal{L}_{ideology}$):] The same triplet-loss objective used by the POLITICS model (Section~\ref{sec:transformer-baselines}, Equation~\ref{eqn:triplet_loss}), which positions articles from ideologically similar sources closer in representation space than ideologically distinct articles. We adopt cosine distance for $d(\cdot,\cdot)$ and set the margin $\delta = 1$. Unlike \citet{liu2022politics}, who combine story-level and ideology-level contrastive objectives, we follow \citet{baly2020we} and retain only the ideology objective; \citet{liu2022politics} themselves report that the story objective degrades ideology classification when applied jointly.
    
    \item[\textbf{Thematic Classification Loss} ($\mathcal{L}_{themes}$):] Multi-label binary cross-entropy for predicting GDELT theme annotations, enabling the model to learn associations between ideological positions and topic coverage patterns:
    
    \begin{equation}
    \mathcal{L}_{themes} = -\frac{1}{K} \sum_{k=1}^{K} \left[ y_k \log(\hat{y}_k) + (1-y_k) \log(1-\hat{y}_k) \right] \label{eq:theme-loss}
    \end{equation}
    
    where $K$ is the number of theme categories, $y_k$ is the binary ground truth for theme $k$, and $\hat{y}_k$ is the predicted probability.
    
    \item[\textbf{Tonal Regression Loss} ($\mathcal{L}_{tone}$):] Mean squared error for predicting GDELT sentiment scores, allowing the model to capture emotional framing patterns:
    
    \begin{equation}
    \mathcal{L}_{tone} = \frac{1}{N} \sum_{i=1}^{N} (s_i - \hat{s}_i)^2 \label{eq:tone-loss}
    \end{equation}
    
    where $s_i$ is the true tone score and $\hat{s}_i$ is the predicted score for article $i$.

    \item[\textbf{Masked Language Modeling Loss} ($\mathcal{L}_{MLM}$):] The standard masked language modeling objective from RoBERTa, which preserves the model's general language understanding capabilities during continued pre-training.
\end{description}

This multi-task approach follows the principle that shared representations beneficial for auxiliary tasks (theme classification and tone prediction) will also improve the primary task (ideological classification) by capturing more comprehensive discourse patterns. The first-principles argument for why \emph{topical coverage} should carry ideological signal is concrete: outlets with different ideological orientations systematically emphasize different mixes of topics: immigration is framed in terms of border security in some outlets and humanitarian crisis in others, healthcare is framed around cost in some and access in others, and so on. By forcing the encoder to predict GDELT theme labels explicitly during pre-training, we inject this topic-distribution signal directly into the encoder rather than hoping the downstream classifier discovers it from raw token statistics.

\paragraph{Why auxiliary-task supervision rather than feature concatenation.} An obvious alternative would be to simply concatenate GDELT theme and tone features as additional inputs at the classifier layer. We choose auxiliary-task supervision instead because it regularizes the encoder's \emph{internal} representation space rather than augmenting the input layer: the metalinguistic structure becomes part of how the encoder represents text in general, and remains available to downstream tasks even when GDELT features are not present at inference time. Concatenating features at the input layer, by contrast, would couple downstream performance to the continued availability of GDELT metadata and would not change how the encoder reads articles that lack such metadata.

\subsection{Experimental Results}
\label{sec:continued-pretraining-results}

We employed the RoBERTa-base architecture \citep{liu2019roberta} as our foundation model, initializing from the standard pre-trained checkpoint before applying our continued pre-training regime for 2 epochs over the GDELT corpus. We also considered only the 2000 most frequent GDELT categories, and the triplet loss margin $\delta$ is set to 1. Specific pre-training settings can be found in Table~\ref{tab:pretraining_config}.

We conducted experiments varying two key factors: the number of triplet samples per batch (1, 16, or 32) and the inclusion of metalinguistic objectives (theme prediction $T_h$, tone regression $T_o$, or both). Triplet generation used in-batch positives and negatives drawn by random sampling: for each anchor article in a batch, positives (same ideology label) and negatives (different ideology label) were sampled uniformly at random from the other articles in the same batch, with no explicit hard-negative mining or curriculum.

Following continued pre-training, we fine-tuned all models on the AllSidesV2 Media training split and evaluated on the test split. Each configuration was trained for two epochs, with Table~\ref{tab:pretraining-combined} reporting the best performance achieved across epochs.

\begin{table}[ht]
\centering
\footnotesize
\begin{tabular}{@{}lcccccccc@{}}
\toprule
\textbf{Experiment} & \multicolumn{4}{c}{\textbf{Subset (1.6M, Complete GDELT)}} & \multicolumn{4}{c}{\textbf{Full (4.5M, 35\% GDELT)}} \\
 & F1-Macro & Left & Center & Right & F1-Macro & Left & Center & Right \\
\midrule
$T_{id}(1)$          & 50.12 & 69.28 & 18.09 & 63.00 & 56.98 & 73.82 & 33.96 & 63.55 \\
$T_{id}(1) + T_h$    & \cellcolor{blue!15}52.02 & \cellcolor{blue!15}71.61 & \cellcolor{blue!15}\textbf{33.41} & \cellcolor{blue!15}51.03 & 51.40 & 68.17 & 21.81 & 64.21 \\
$T_{id}(1) + T_o$    & 53.39 & 71.65 & 28.31 & 60.20 & 55.86 & 72.46 & 31.17 & 63.93 \\
$T_{id}(1) + T_h + T_o$ & \textbf{56.56} & 72.58 & 30.39 & 66.71 & 60.78 & 73.05 & 39.48 & 69.82 \\
\midrule
$T_{id}(16)$         & 50.06 & 72.07 & 20.53 & 57.59 & 52.88 & 68.36 & 22.02 & 68.26 \\
$T_{id}(16) + T_h$   & 50.10 & 67.33 & 16.23 & 66.74 & \cellcolor{blue!15}60.55 & \cellcolor{blue!15}74.11 & \cellcolor{blue!15}38.22 & \cellcolor{blue!15}69.31 \\
$T_{id}(16) + T_o$   & 53.30 & 71.99 & 22.08 & 65.83 & 53.76 & 74.11 & 19.71 & 67.45 \\
$T_{id}(16) + T_h + T_o$ & \textbf{56.56} & 72.58 & 30.39 & 66.71 & 60.78 & 73.05 & 39.48 & 69.82 \\
\midrule
$T_{id}(32)$         & 55.41 & \textbf{73.66} & 28.34 & 64.22 & 60.15 & 73.95 & 35.88 & 70.62 \\
$T_{id}(32) + T_h$   & 46.47 & 66.05 & 03.83 & \textbf{69.52} & \textbf{62.19} & \textbf{75.97} & 39.83 & \textbf{70.76} \\
$T_{id}(32) + T_o$   & 48.74 & 67.21 & 12.83 & 66.18 & \cellcolor{blue!15}57.18 & \cellcolor{blue!15}72.75 & \cellcolor{blue!15}34.49 & \cellcolor{blue!15}64.30 \\
$T_{id}(32) + T_h + T_o$ & \cellcolor{blue!15}55.59 & \cellcolor{blue!15}72.73 & \cellcolor{blue!15}29.27 & \cellcolor{blue!15}64.76 & 59.33 & 72.66 & \textbf{42.28} & 63.06 \\
\bottomrule
\end{tabular}
\caption{Continued pre-training results on both GDELT limited subset and full \textsc{BigNewsBLN} dataset. Best results per column in bold. \colorbox{blue!15}{Highlighted} cells indicate configurations where epoch~1 outperformed epoch~2. All models fine-tuned on AllSidesV2 Media train and evaluated on test split.}
\label{tab:pretraining-combined}
\end{table}

Table~\ref{tab:pretraining-combined} presents results from pre-training on both the 1.6 million article subset with complete GDELT coverage, and the full 4.5 million article \textsc{BigNewsBLN} corpus, where metalinguistic objectives were applied only to the articles with available GDELT metadata. Our most effective configuration, $T_{32} + T_h$, achieved 62.19 F1-Macro on the full dataset, outperforming the POLITICS baseline (60.60) and demonstrating the value of combining intensive triplet training with theme prediction. The full results for each configuration across both epochs can be found in Appendix~\ref{sec:appendix-full-experiments}.

\paragraph{Subset-vs-full inversions.} Several configurations in Table~\ref{tab:pretraining-combined} get \emph{worse} when given more pre-training data; most notably $T_{id}(1)+T_h$, which falls from 52.02 on the 1.6M-article subset (where every article has GDELT metadata) to 51.40 on the full 4.5M-article corpus (where only 35\% of articles have GDELT metadata). The most plausible mechanism is auxiliary-signal density: on the full corpus, the auxiliary heads receive supervision on only $\sim$35\% of training examples, gradient noise on the auxiliary objectives rises, and the effective contribution of the metalinguistic loss is down-weighted at the same nominal loss-mixing weight. Configurations that lean more heavily on the auxiliary signal (small triplet sample, theme-only) are exactly the ones most exposed to this signal-density drop, and they are exactly the ones that regress.

\paragraph{Epoch-1 outperformance.} Several ``best'' cells in Table~\ref{tab:pretraining-combined} are reached at epoch~1 rather than epoch~2 (highlighted in blue). This is one of the more interesting empirical patterns in this chapter and warrants further investigation. We discuss it in greater detail in Appendix~\ref{sec:appendix-full-experiments}, but it is most likely attributable to incomplete GDELT metadata coverage interacting with random triplet sampling: under sparse auxiliary supervision, a single epoch is enough for the auxiliary heads to influence the encoder representation, while a second epoch begins to over-fit to the noisy GDELT-derived targets without compensating gains on the ideology objective.

These findings extend prior work \citep{smirnova2017ideology,bhatia2018topic} demonstrating the utility of metalinguistic cues such as sentiment and emotional tone in political ideology detection. Our results illustrate that explicitly incorporating such signals as auxiliary training objectives can yield substantial performance improvements for article-level political ideology classification. The best performing models (both \href{https://huggingface.co/dragonslayer631/roberta_metalinguistic_features_triplet_loss}{pre-trained checkpoints} and \href{https://huggingface.co/dragonslayer631/ideology_classifier_finetuned}{fine-tuned versions} on the AllSidesV2 dataset) are publicly available on HuggingFace.\footnote{Models available under the Apache 2.0 license.}

% TODO: Add error analysis to understand where the improved model still fails
% TODO: Add qualitative analysis of specific examples where metalinguistic features help
