
\section{Problem Formulation and Technical Approach}
\label{sec:problem-formulation}

Existing approaches to political bias classification rely primarily on semantic content analysis, either through fine-tuned transformer models or specialized contrastive learning frameworks \citep{baly2020we, liu2022politics}. However, political ideology in news articles manifests not only through explicit content but also through metalinguistic markers such as topic selection, framing strategies, and emotional tone. Cognitive science research has long established that metalinguistic awareness---the ability to reflect upon and manipulate the structural features of language---is foundational to reading comprehension \citep{tunmer1984development, nagy2007metalinguistic, Zipke10092007}. In NLP, prior work has demonstrated that sentiment and thematic content carry ideological signal: \citet{smirnova2017ideology} used lexicon-based sentiment analysis to trace ideological shifts in news coverage, while \citet{bhatia2018topic} showed that topic-specific sentiment polarities identify ideology in both speeches and news articles. At the source level, \citet{ronnback2025biasoutlets} demonstrated that tone and topic features classify outlet-level political bias at 76\% accuracy against a 45\% majority-class baseline. Despite this evidence, these higher-level discourse features remain largely unexploited in article-level computational approaches.

We formulate the integration of metalinguistic features as a multi-task learning problem where the primary political bias classification objective is augmented with auxiliary tasks that capture complementary aspects of political discourse. The primary task is political ideology classification using contrastive triplet-loss objectives. This is supplemented by two auxiliary tasks: multi-label topic classification capturing thematic focus, and regression of sentiment polarity reflecting tonal characteristics. This formulation enables the model to learn representations that are sensitive to both explicit ideological content and implicit stylistic markers that correlate with political orientation.

